\clearpage
\section{Algebraic Graph Theory: Auxiliary Topics}

This chapter is a graduate-level tour of the corner of algebraic
combinatorics that interacts most strongly with graph theory.
Each section pairs an algebraic construction (a polynomial, a Hopf
algebra, a Chow ring) with a concrete graph-theoretic payoff
(real-rootedness of the matchings polynomial, the chromatic symmetric
function, log-concavity of the chromatic polynomial, rapid mixing of
matroid Markov chains). The chapter is largely independent of
Chapter~11 but freely cites spectral results from \S 11.4.

\subsubsection*{Roadmap and graph-theoretic motivation}

\begin{itemize}
  \item \S 12.1 \textbf{Matchings polynomial / Heilmann--Lieb.} A
        graph polynomial whose roots are always real. Generalises the
        adjacency-spectrum theorem from forests to all graphs;
        equivalent to the absence of phase transition in the
        monomer--dimer model.
  \item \S 12.2 \textbf{Equitable partitions / interlacing.} Symmetry
        groups of a graph induce equitable partitions; their quotient
        matrices give explicit eigenvalues. Yields the Hoffman /
        Haemers ratio bounds for $\alpha(G)$ and $\chi(G)$.
  \item \S 12.3 \textbf{Distance-regular graphs / association schemes.}
        Algebraic generalisation of strongly regular graphs.
        Underlies Delsarte's LP bound for codes (Hamming scheme) and
        Erd\H{o}s--Ko--Rado-type results (Johnson scheme).
  \item \S 12.4 \textbf{Cayley graphs.} Eigenvalues of $\mathrm{Cay}(\Gamma, S)$
        from characters of $\Gamma$; explicit Ramanujan graph
        constructions; the spectral source for nearly all
        deterministic expanders.
  \item \S 12.5 \textbf{Tutte polynomial / matroids.} Universal
        deletion--contraction graph invariant; specialises to
        chromatic, flow, reliability, Ising, Jones polynomials. The
        matroid framework abstracts cycle/bond duality.
  \item \S 12.6 \textbf{Heaps of pieces.} Walk generating functions on
        graphs as continued fractions; combinatorial proof of
        Heilmann--Lieb; bridge to orthogonal polynomial theory.
  \item \S 12.7 \textbf{Ihara zeta function.} Graph Riemann zeta whose
        ``Riemann hypothesis'' equals the Ramanujan property;
        Hashimoto non-backtracking matrix is the foundation of modern
        spectral community detection.
  \item \S 12.8 \textbf{Combinatorial Hopf algebras.} Schmitt's graph
        Hopf algebra recovers Stanley's chromatic symmetric function
        via the Aguiar--Bergeron--Sottile universal property;
        Stanley--Stembridge conjecture and Hikita's recent proof.
  \item \S 12.9 \textbf{Hodge theory of matroids.} Adiprasito--Huh--Katz
        prove the chromatic polynomial coefficients are log-concave by
        importing K\"ahler-package machinery into combinatorics.
  \item \S 12.10 \textbf{Permanent and \#P.} Counting perfect matchings
        is \#P-complete; Jerrum--Sinclair--Vigoda FPRAS uses Markov
        chains on the matching lattice; holographic algorithms map
        graph problems to permanents.
  \item \S 12.11 \textbf{Weisfeiler--Leman and graph isomorphism.}
        Color refinement, the $k$-WL hierarchy, the
        Cai--F\"urer--Immerman counterexample, Babai's quasipolynomial
        algorithm, and the connection to graph neural network
        expressivity.
  \item \S 12.12 \textbf{Cluster algebras from quivers.} Quivers as
        directed graphs; mutation and cluster variables; cluster
        algebras of finite type and the Cartan--Killing classification;
        triangulations of surfaces.
  \item \S 12.13 \textbf{Chip-firing, sandpiles, and Riemann--Roch for
        graphs.} The sandpile (Picard) group of a graph; Baker--Norine
        Riemann--Roch theorem; explicit graph-theoretic analogues of
        algebraic-geometric intersection theory.
  \item \S 12.14 \textbf{Random walks via representation theory.}
        Diaconis--Shahshahani upper-bound lemma; cutoff phenomenon;
        explicit mixing-time bounds for the riffle shuffle, the
        hypercube, and Cayley graphs of the symmetric group.
  \item \S 12.15 \textbf{Electrical networks and effective
        resistance.} Resistor networks dual to random walks; the
        spanning-tree-counting / commute-time / hitting-time identities
        of Chandra--Raghavan--Ruzzo--Smolensky; Schramm--Lov\'asz on
        circle packings.
  \item \S 12.16 \textbf{Crystal graphs and the plactic monoid.}
        Kashiwara crystals as edge-labelled graphs whose connected
        components encode irreducible $\mathfrak{gl}_n$-representations;
        Knuth equivalence and the plactic structure on words.
\end{itemize}

\paragraph{References (canonical).}
Godsil--Royle, \emph{Algebraic Graph Theory} (2001);
Godsil, \emph{Algebraic Combinatorics} (1993);
Stanley, \emph{Algebraic Combinatorics: Walks, Trees, Tableaux} (UTM,
2013); Garsia--Egecioglu, \emph{Lectures in Algebraic Combinatorics}
(2020); Brouwer--Haemers, \emph{Spectra of Graphs} (2012);
Brouwer--Cohen--Neumaier, \emph{Distance-Regular Graphs} (1989);
Bannai--Ito, \emph{Algebraic Combinatorics I} (1984); Oxley,
\emph{Matroid Theory} (2nd ed., 2011); Diaconis,
\emph{Group Representations in Probability and Statistics} (1988);
Lyons--Peres, \emph{Probability on Trees and Networks} (2016);
Bump, \emph{Lie Groups} (2nd ed., 2013) for crystal bases.
Selected research papers and theses are cited inline throughout.

%====================================================================
\subsection{The Matchings Polynomial and the Heilmann--Lieb Theorem}
%====================================================================

The adjacency-matrix characteristic polynomial $\phi(G, x) = \det(xI - A(G))$
encodes the spectrum of $G$, but it is highly non-trivial to relate to
the combinatorial structure of $G$. The \emph{matchings polynomial}
plays a parallel role for matchings and turns out to coincide with
$\phi(G, x)$ when $G$ is a forest, while always remaining a polynomial
with real roots only.

\subsubsection{Definition and Basic Recurrences}

\begin{definition}[Matchings polynomial]\label{def:matchpoly}
Let $G$ be a graph on $n$ vertices and let $p(G, k)$ denote the number
of $k$-matchings (sets of $k$ pairwise non-adjacent edges). The
\textbf{matchings polynomial} of $G$ is
\[
  \mu(G, x) \;=\; \sum_{k \ge 0} (-1)^k\, p(G, k)\, x^{n - 2k}.
\]
By convention, $p(G, 0) = 1$ and $\mu(\emptyset, x) = 1$.
\end{definition}

\begin{example}[Small graphs]
\mbox{}
\begin{itemize}
  \item $\mu(K_2, x) = x^2 - 1$.
  \item $\mu(P_3, x) = x^3 - 2x$.
  \item $\mu(C_3, x) = x^3 - 3x$ (three edges, no $2$-matching).
  \item $\mu(C_4, x) = x^4 - 4x^2 + 2$ (two $2$-matchings: opposite-edge pairs).
\end{itemize}
\end{example}

\begin{theorem}[Edge and vertex recurrences]\label{thm:matchrec}
For any edge $e = uv$ and any vertex $u$ of $G$,
\begin{align*}
  \mu(G, x) &\;=\; \mu(G \setminus e, x) \;-\; \mu(G \setminus \{u, v\}, x), \\
  \mu(G, x) &\;=\; x\,\mu(G \setminus u, x) \;-\; \sum_{v \sim u} \mu(G \setminus \{u, v\}, x).
\end{align*}
\end{theorem}

\begin{proof}
For the first identity, partition $k$-matchings of $G$ by whether
$e \in M$. Matchings not containing $e$ are $k$-matchings of $G \setminus e$,
contributing $\mu(G \setminus e, x)$. Matchings containing $e$ are
$(k - 1)$-matchings of $G \setminus \{u, v\}$, contributing
$x^2 \mu(G \setminus \{u, v\}, x) / x^2 = \mu(G \setminus \{u, v\}, x)$
once the sign is tracked. The vertex recurrence is the edge recurrence
summed over all edges incident to $u$.
\end{proof}

\begin{proposition}[Matching polynomial of a forest]\label{prop:matchforest}
If $T$ is a forest, then $\mu(T, x) = \phi(T, x) = \det(xI - A(T))$.
\end{proposition}

\begin{proof}
The Leibniz formula expresses $\det(xI - A)$ as a sum over permutations
$\sigma$, which decompose into cycles. In a forest the only cycles
admissible to a non-zero contribution are fixed points (giving an $x$)
and $2$-cycles (giving $-1$ via an edge). The non-fixed-point support
of $\sigma$ thus selects a matching $M$, and the contribution is
$(-1)^{|M|} x^{n - 2|M|}$. Summing matches Definition~\ref{def:matchpoly}.
\end{proof}

\subsubsection{The Path-Tree Construction}

The fundamental insight of Godsil and Heilmann--Lieb is that for any
graph $G$ with a distinguished vertex $u$, one can build a tree
$T(G, u)$ whose characteristic polynomial controls the matchings
polynomial of $G$.

\begin{definition}[Path-tree]\label{def:pathtree}
Let $G$ be a graph and $u \in V(G)$. The \textbf{path-tree}
$T(G, u)$ is the rooted tree whose vertices are the non-backtracking
walks $u = v_0 \to v_1 \to \cdots \to v_k$ in $G$ starting at $u$,
and whose edges connect each walk of length $k$ to its $(k+1)$-step
extensions in $G$.
\end{definition}

\begin{theorem}[Godsil's path-tree identity]\label{thm:pathtree}
For any vertex $u$ of $G$,
\[
  \frac{\mu(G \setminus u, x)}{\mu(G, x)}
  \;=\; \frac{\mu(T(G, u) \setminus u, x)}{\mu(T(G, u), x)}.
\]
In words, the ratio is identical to the corresponding ratio for the
characteristic polynomials of the path-tree.
\end{theorem}

\begin{corollary}[Reduction to a tree]\label{cor:rootreduce}
Every root of $\mu(G, x)$ is a root of $\phi(T(G, u), x)$ for any
choice of $u$. Since $T(G, u)$ is a tree,
$\phi(T(G, u), x)$ has only real roots; therefore so does
$\mu(G, x)$.
\end{corollary}

\subsubsection{The Heilmann--Lieb Theorem}

\begin{theorem}[Heilmann--Lieb, 1972]\label{thm:hl}
Let $G$ be a graph with maximum degree $\Delta$. Then all roots of
$\mu(G, x)$ are real, and they lie in the open interval
$(-2\sqrt{\Delta - 1},\; 2\sqrt{\Delta - 1})$.
\end{theorem}

\begin{proof}[Proof sketch]
Reality follows from Corollary~\ref{cor:rootreduce}. For the bound,
observe that the path-tree $T(G, u)$ has maximum degree at most
$\Delta$, so its adjacency spectrum is contained in
$[-2\sqrt{\Delta - 1}, 2\sqrt{\Delta - 1}]$ by the Alon--Boppana
analysis of the universal cover. Strict inclusion in the open interval
follows from the fact that $T(G, u)$ is finite and the spectral radius
of any finite tree of bounded degree $\Delta$ is strictly less than
$2\sqrt{\Delta - 1}$.
\end{proof}

\begin{remark}[Statistical-physics interpretation]
The matchings polynomial is, up to substitution, the partition function
of the \emph{monomer--dimer model} on $G$:
$Z_G(\lambda) = \sum_M \lambda^{|M|}$ where the sum runs over all
matchings. Theorem~\ref{thm:hl} implies $Z_G(\lambda) > 0$ for all real
$\lambda \ge 0$, so the monomer--dimer model has \emph{no phase transition}
on any finite graph.
\end{remark}

\subsubsection{Special Cases and Classical Polynomials}

\begin{theorem}[Matching polynomials of named graphs]
\mbox{}
\begin{itemize}
  \item $\mu(K_n, x)$ is (up to scaling) the Hermite polynomial $H_n(x)$.
  \item $\mu(P_n, x)$ is the Chebyshev polynomial of the second kind
        $U_n(x/2)$.
  \item $\mu(K_{m, n}, x)$ is (up to scaling) a Laguerre polynomial.
\end{itemize}
\end{theorem}

\begin{example}[Matching polynomial of $K_4$]
$K_4$ has six edges and three perfect matchings. Direct expansion:
$p(K_4, 0) = 1$, $p(K_4, 1) = 6$, $p(K_4, 2) = 3$. Hence
\[
  \mu(K_4, x) \;=\; x^4 - 6 x^2 + 3.
\]
The roots are $\pm \sqrt{3 \pm \sqrt{6}}$, all real and inside the
Heilmann--Lieb interval $(-2\sqrt{2}, 2\sqrt{2}) \approx (-2.83, 2.83)$.
\end{example}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute $\mu(C_5, x)$ and $\mu(C_6, x)$ from the recurrence
$\mu(C_n, x) = \mu(P_n, x) - \mu(P_{n-2}, x)$, which follows from
Theorem~\ref{thm:matchrec} applied to a single edge of $C_n$.
\end{exercise}

\begin{exercise}
Verify Proposition~\ref{prop:matchforest} for the path $P_4$ by
computing both $\mu(P_4, x)$ from the matching definition and
$\det(xI - A(P_4))$ from the adjacency matrix.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Prove $\mu(G, x) = \mu(G \setminus e, x) - \mu(G \setminus \{u, v\}, x)$
in detail by induction on $|E(G)|$, taking care of the edge-case
$|M| = 0$.
\end{exercise}

\begin{exercise}
Show that $\mu(G \cup H, x) = \mu(G, x) \cdot \mu(H, x)$ for the
disjoint union of $G$ and $H$. Use this to compute the matchings
polynomial of the $5$-cube $Q_5$ from $\mu(Q_4, x)$ and $\mu(K_2, x)$
(Hint: $Q_5 = Q_4 \,\square\, K_2$ requires more work; treat it as
two copies of $Q_4$ joined by a matching).
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove the path-tree identity (Theorem~\ref{thm:pathtree}) by
applying the vertex recurrence simultaneously to $G$ and to
$T(G, u)$ at the root, then inducting on the depth of expansion.
\end{exercise}

\begin{exercise}
Use the Heilmann--Lieb theorem to deduce that the partition function
$Z_G(\lambda) = \sum_M \lambda^{|M|}$ has no real positive zeros for
$\lambda > 0$. Conclude that $\log Z_G(\lambda)$ is real-analytic on
$(0, \infty)$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the path-tree $T(C_4, v)$ for any vertex $v$ of the $4$-cycle
$C_4$. Verify that $T(C_4, v)$ is the infinite $2$-regular tree
truncated to depth at most $4$, and write out the first few terms of
its characteristic polynomial.
\end{exercise}

\begin{exercise}
Mark the roots of $\mu(K_4, x)$ on the real line and shade the
Heilmann--Lieb interval $(-2\sqrt{2}, 2\sqrt{2})$ to confirm strict
containment. Compare with the actual adjacency spectrum
$\{3, -1, -1, -1\}$ of $K_4$ (which lies on $\mathbb{Z}$, not on the
matching polynomial roots).
\end{exercise}

%====================================================================
\subsection{Equitable Partitions and Quotient Interlacing}
%====================================================================

Equitable partitions provide a systematic way to extract eigenvalues
from a graph using its symmetries. Every automorphism orbit gives an
equitable partition, but many graphs admit equitable partitions with
no automorphism behind them.

\subsubsection{Definition and Basic Theory}

\begin{definition}[Equitable partition]\label{def:equitable}
A partition $\pi = (V_1, V_2, \ldots, V_k)$ of $V(G)$ is
\textbf{equitable} if there exist non-negative integers
$b_{ij}$ ($1 \le i, j \le k$) such that every vertex of $V_i$ has
exactly $b_{ij}$ neighbors in $V_j$. The $k \times k$ matrix
$B = (b_{ij})$ is the \textbf{quotient matrix} of the partition.
\end{definition}

\begin{example}[Trivial partitions]
\mbox{}
\begin{itemize}
  \item The discrete partition into singletons is always equitable;
        its quotient matrix is the adjacency matrix $A(G)$.
  \item For a $k$-regular graph, the trivial $1$-cell partition
        $\pi = (V)$ is equitable with $B = (k)$.
  \item For a bipartite graph $G = (A \cup B, E)$ where every vertex
        of $A$ has degree $d_A$ and every vertex of $B$ has degree
        $d_B$, the partition $(A, B)$ is equitable with
        $B = \big(\begin{smallmatrix} 0 & d_A \\ d_B & 0 \end{smallmatrix}\big)$.
\end{itemize}
\end{example}

\begin{theorem}[Spectrum of the quotient]\label{thm:quotientspec}
If $\pi$ is an equitable partition of $G$ with quotient matrix $B$,
then every eigenvalue of $B$ is also an eigenvalue of $A(G)$.
Specifically, if $By = \lambda y$ then the vector $x \in \mathbb R^V$
defined by $x_v = y_i$ whenever $v \in V_i$ satisfies $A x = \lambda x$.
\end{theorem}

\begin{proof}
Let $S \in \{0, 1\}^{V \times k}$ be the characteristic matrix of
$\pi$ (column $i$ is the indicator of $V_i$). Equitability means
$AS = SB$. Hence if $By = \lambda y$, then
$A(Sy) = S(By) = \lambda(Sy)$. The lifted vector $Sy$ is non-zero
because the columns of $S$ are linearly independent.
\end{proof}

\subsubsection{Cauchy Interlacing for Quotients}

\begin{theorem}[Cauchy interlacing for quotients]\label{thm:quotinter}
Let $A$ be a real symmetric $n \times n$ matrix with eigenvalues
$\lambda_1 \ge \cdots \ge \lambda_n$, and let $\pi$ be any partition
(not necessarily equitable) with characteristic matrix $S$. Define the
\emph{normalized quotient}
$\hat B = (S^\top S)^{-1} S^\top A S$, a $k \times k$ matrix with
eigenvalues $\beta_1 \ge \cdots \ge \beta_k$. Then
\[
  \lambda_i \;\ge\; \beta_i \;\ge\; \lambda_{n - k + i}
  \qquad (1 \le i \le k).
\]
\end{theorem}

\begin{proof}
Apply the Courant--Fischer min-max characterization
(Theorem~\ref{thm:courantfischer}) to $A$ and to $\hat B = S^\top A S$
on the $k$-dimensional column span of $S$. The Rayleigh quotient on
this $k$-subspace lies between the top-$k$ and bottom-$k$ eigenvalues
of $A$.
\end{proof}

\begin{corollary}[Hoffman-style ratio bound; Haemers, 1995]\label{cor:haemers}
Let $G$ be a $k$-regular graph with smallest adjacency eigenvalue
$\lambda_n$. Then
\[
  \alpha(G) \;\le\; \frac{-\lambda_n}{k - \lambda_n} \cdot n.
\]
\end{corollary}

\begin{proof}
Let $S \subseteq V(G)$ be a maximum independent set, and consider the
partition $(S, V \setminus S)$. Compute the normalized quotient: the
$2 \times 2$ matrix has trace $2 \cdot \frac{e(S, V \setminus S)}{|S|}$
and determinant relating to the densities. Apply
Theorem~\ref{thm:quotinter} and solve.
\end{proof}

\subsubsection{Walk-Regularity and Distance Partitions}

\begin{definition}[Walk-regular]
A graph $G$ is \textbf{walk-regular} if for every $k \ge 0$, the
diagonal $(A^k)_{vv}$ is the same for all $v$. Equivalently, the number
of closed walks of length $k$ at $v$ does not depend on $v$.
\end{definition}

\begin{theorem}[Vertex-transitive implies walk-regular]
Every vertex-transitive graph is walk-regular.
\end{theorem}

\begin{definition}[Distance partition]
For a vertex $v$ in a connected graph $G$, the \textbf{distance
partition from $v$} is
$\pi_v = (G_0(v), G_1(v), G_2(v), \ldots, G_d(v))$, where $G_i(v)$ is
the set of vertices at distance exactly $i$ from $v$.
\end{definition}

\begin{theorem}[Distance partition of a distance-regular graph]\label{thm:drgequit}
A connected graph $G$ is distance-regular (Section~\ref{subsec:drg}
below) iff for every vertex $v$, the distance partition $\pi_v$ is
equitable with the same quotient matrix.
\end{theorem}

This sets up Section~\ref{subsec:drg}, where the quotient matrix is
the canonical tridiagonal intersection matrix.

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Show that the orbit partition of any subgroup of $\mathrm{Aut}(G)$ is
equitable.
\end{exercise}

\begin{exercise}
Find an equitable partition of the Petersen graph into two parts
(outer pentagon and inner pentagram). Write its $2 \times 2$ quotient
matrix and verify both eigenvalues appear in the Petersen spectrum
$\{3, 1^{(5)}, -2^{(4)}\}$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Prove Theorem~\ref{thm:quotientspec} from the matrix identity
$AS = SB$. Show further that the $k$ eigenvalues of $B$ are exactly
the eigenvalues of $A$ whose eigenvectors are constant on each cell
of $\pi$.
\end{exercise}

\begin{exercise}
Apply the Haemers ratio bound (Corollary~\ref{cor:haemers}) to the
Petersen graph and to the Kneser graph $K(5, 2)$ (which is the
Petersen graph). Compare with the exact independence number $\alpha(P) = 4$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove that a connected graph is distance-regular iff every distance
partition $\pi_v$ ($v \in V$) is equitable with the \emph{same}
quotient matrix (Theorem~\ref{thm:drgequit}).
\end{exercise}

\begin{exercise}
Let $G$ be a graph with an equitable partition $\pi$. Show that the
spectrum of $G$ decomposes as $\sigma(B) \cup \sigma(A_{\pi^\perp})$,
where $A_{\pi^\perp}$ is $A$ restricted to vectors that sum to zero on
each cell. Conclude that all ``new'' eigenvalues of $G$ relative to
$B$ have eigenvectors orthogonal to the constant-on-cells subspace.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the cycle $C_6$ and find an equitable partition into three cells
of size $2$ each (alternating cells). Write the $3 \times 3$ quotient
matrix and verify that its eigenvalues $\{2, -1, -1\}$ appear among
the cycle spectrum $\{2, 1, -1, -2, -1, 1\}$.
\end{exercise}

\begin{exercise}
Sketch the distance partition of the cube $Q_3$ from a corner vertex.
Verify that it is equitable into four cells of sizes $1, 3, 3, 1$, and
write its tridiagonal quotient matrix.
\end{exercise}

%====================================================================
\subsection{Distance-Regular Graphs and Association Schemes}\label{subsec:drg}
%====================================================================

Distance-regular graphs (DRGs) sit at the algebraic core of extremal
combinatorics: they include all strongly regular graphs (Section 11.4),
the Hamming and Johnson graphs (which underlie coding theory and
the Erd\H{o}s--Ko--Rado problem), and the foundational examples of
finite geometry. Their unified algebraic theory is the language of
\emph{association schemes}.

\subsubsection{Distance-Regular Graphs}

\begin{definition}[Distance-regular]\label{def:drg}
A connected graph $G$ of diameter $d$ is \textbf{distance-regular}
if there exist integers $b_0, b_1, \ldots, b_{d-1}$ and
$c_1, c_2, \ldots, c_d$ such that, for every pair $u, v$ at distance
$i$ in $G$,
\[
  c_i \;=\; |N(v) \cap G_{i-1}(u)|,
  \qquad
  b_i \;=\; |N(v) \cap G_{i+1}(u)|,
\]
i.e.\ the neighbours of $v$ counted by their distance from $u$ depend
only on $i$. The sequence
\[
  \iota(G) \;=\; \{b_0, b_1, \ldots, b_{d-1};\; c_1, c_2, \ldots, c_d\}
\]
is the \textbf{intersection array}. The valency is $b_0 = k$.
\end{definition}

\begin{example}[Familiar distance-regular graphs]
\mbox{}
\begin{itemize}
  \item $K_n$ has intersection array $\{n - 1; 1\}$ (diameter $1$).
  \item $C_n$ ($n \ge 5$) has intersection array
        $\{2, 1, 1, \ldots, 1; 1, 1, \ldots, 1, c_d\}$, where the last
        $c_d \in \{1, 2\}$ depending on parity.
  \item The Petersen graph has $\iota = \{3, 2; 1, 1\}$ (diameter $2$).
  \item The Hamming graph $H(n, q)$ (vertices: words in $[q]^n$;
        edges: words differing in one coordinate) is distance-regular
        with $b_i = (n - i)(q - 1)$ and $c_i = i$.
  \item The Johnson graph $J(n, k)$ (vertices: $k$-subsets of $[n]$;
        edges: subsets sharing $k - 1$ elements) is distance-regular
        with $b_i = (k - i)(n - k - i)$ and $c_i = i^2$.
\end{itemize}
\end{example}

\begin{theorem}[Tridiagonal quotient]\label{thm:drgquot}
Let $G$ be distance-regular with intersection array
$\{b_0, \ldots, b_{d-1}; c_1, \ldots, c_d\}$. Set $a_i = k - b_i - c_i$
for $0 \le i \le d$ (where $b_d = c_0 = 0$). The quotient matrix of
the distance partition $\pi_v$ from any vertex $v$ is
\[
  L_1 \;=\;
  \begin{pmatrix}
    a_0 & b_0 & 0 & \cdots & 0 \\
    c_1 & a_1 & b_1 & \cdots & 0 \\
    0 & c_2 & a_2 & \cdots & 0 \\
    \vdots & & \ddots & \ddots & \vdots \\
    0 & 0 & \cdots & c_d & a_d
  \end{pmatrix}.
\]
By Theorem~\ref{thm:quotientspec}, every eigenvalue of $L_1$ is an
eigenvalue of $A(G)$; in fact $G$ has \emph{exactly} $d + 1$ distinct
adjacency eigenvalues, the eigenvalues of $L_1$.
\end{theorem}

\subsubsection{Association Schemes}

\begin{definition}[Association scheme]\label{def:scheme}
A (symmetric) \textbf{association scheme} with $d$ classes on a finite
set $X$ is a partition $\{R_0, R_1, \ldots, R_d\}$ of $X \times X$ such
that, with $A_i \in \{0, 1\}^{X \times X}$ the indicator matrix of
$R_i$:
\begin{enumerate}
  \item $R_0 = \{(x, x) : x \in X\}$, so $A_0 = I$.
  \item Each $R_i$ is symmetric: $A_i^\top = A_i$.
  \item $\sum_{i = 0}^{d} A_i = J$ (the all-ones matrix).
  \item There exist non-negative integers $p_{ij}^k$ with
        $A_i A_j = \sum_{k = 0}^{d} p_{ij}^k A_k$.
\end{enumerate}
The numbers $p_{ij}^k$ are the \textbf{intersection numbers} of the
scheme.
\end{definition}

\begin{theorem}[Bose--Mesner algebra]\label{thm:bosemesner}
The complex span of $A_0, A_1, \ldots, A_d$ is a commutative algebra
$\mathfrak A$ under matrix multiplication, called the \textbf{Bose--Mesner
algebra}. Since the $A_i$ are symmetric and commute pairwise,
$\mathfrak A$ has a basis of mutually orthogonal idempotents
$E_0, E_1, \ldots, E_d$ (the \emph{primitive idempotents}). Both
$\{A_i\}$ and $\{E_j\}$ are bases of $\mathfrak A$, related by the
\textbf{first} and \textbf{second eigenmatrices} $P, Q$:
\[
  A_i = \sum_{j} P_{ji} E_j, \qquad E_j = \frac{1}{|X|} \sum_{i} Q_{ij} A_i.
\]
\end{theorem}

\begin{theorem}[Distance-regular = $P$-polynomial]
A distance-regular graph $G$ defines an association scheme on
$X = V(G)$ with $R_i = \{(u, v) : d(u, v) = i\}$. The Bose--Mesner
algebra is generated as a polynomial algebra by $A_1$ (the adjacency
matrix), so distance-regular schemes are exactly the
\emph{$P$-polynomial} schemes.
\end{theorem}

\subsubsection{The Hamming and Johnson Schemes}

\begin{example}[Hamming scheme $H(n, q)$]
$X = [q]^n$; $(x, y) \in R_i$ iff the Hamming distance is $i$. The
graph $H(n, q)$ above is the corresponding distance-regular graph.
The Bose--Mesner algebra is the algebra of symmetric Latin-square
type schemes; the $P$-matrix entries are values of \emph{Krawtchouk
polynomials}.
\end{example}

\begin{example}[Johnson scheme $J(n, k)$]
$X = \binom{[n]}{k}$; $(A, B) \in R_i$ iff $|A \cap B| = k - i$. The
$P$-matrix entries are values of \emph{Hahn polynomials}, and the
LP bounds for codes/cliques in $J(n, k)$ specialize to the
Erd\H{o}s--Ko--Rado theorem.
\end{example}

\begin{theorem}[Delsarte LP bound]\label{thm:delsarte}
Let $\Gamma$ be an association scheme on $X$ with first eigenmatrix
$P$. The maximum size of a \emph{clique} (a subset $C$ such that any
two distinct elements of $C$ are in $R_1$) is bounded by the value of
a linear program in variables $x = (x_0, \ldots, x_d)$ with constraints
$x \ge 0$, $x_0 = 1$, $\sum_i x_i = |C|$, and $Q^\top x \ge 0$.
\end{theorem}

The Delsarte LP gives the best-known general bounds on optimal codes
(in $H(n, q)$), set systems with bounded intersection (in $J(n, k)$),
and packings in spherical caps (continuous analogue).

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Verify that the cycle $C_6$ is distance-regular and compute its
intersection array. Compare with the array of the Petersen graph.
\end{exercise}

\begin{exercise}
Show that every strongly regular graph with parameters
$(n, k, \lambda, \mu)$ is distance-regular of diameter $2$ with
intersection array $\{k, k - 1 - \lambda; 1, \mu\}$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Compute the intersection array of the Hamming graph $H(3, 2) = Q_3$.
Verify directly using Definition~\ref{def:drg}.
\end{exercise}

\begin{exercise}
Verify that for the Johnson scheme $J(5, 2)$, the graph $R_1$ is the
Petersen graph (vertices: $2$-subsets of $[5]$; edges: disjoint
$2$-subsets).
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove Theorem~\ref{thm:bosemesner}: the matrices $A_0, \ldots, A_d$
of an association scheme commute pairwise and span a commutative
algebra. Use the symmetry $A_i^\top = A_i$ and the multiplication rule
$A_i A_j = \sum_k p_{ij}^k A_k$.
\end{exercise}

\begin{exercise}
Use the Delsarte LP bound on the Johnson scheme $J(n, k)$ to recover
the Erd\H{o}s--Ko--Rado theorem: a family of pairwise-intersecting
$k$-subsets of $[n]$ ($n \ge 2k$) has size at most $\binom{n-1}{k-1}$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the Hamming graph $H(2, 3)$ (vertices: $\{0, 1, 2\}^2$; edges:
words differing in one coordinate). Identify it as the rook graph
$3 \times 3$ and compute its spectrum.
\end{exercise}

\begin{exercise}
Sketch the partition $\{R_0, R_1, R_2\}$ of the vertex pairs of $C_5$
by distance. Verify the intersection numbers $p_{ij}^k$ from the
multiplication table $A_i A_j = \sum_k p_{ij}^k A_k$.
\end{exercise}

%====================================================================
\subsection{Cayley Graphs and Vertex-Transitive Graphs}
%====================================================================

Cayley graphs encode group structure as graph structure. Their
eigenvalues are computable from the irreducible characters of the
group, making them the cleanest available source of explicit
$d$-regular graphs with prescribed spectral properties (notably,
expanders).

\subsubsection{Definitions and First Examples}

\begin{definition}[Cayley graph]\label{def:cayley}
Let $\Gamma$ be a finite group and $S \subseteq \Gamma \setminus \{e\}$
a symmetric subset (closed under inversion: $s \in S \Rightarrow s^{-1} \in S$).
The \textbf{Cayley graph} $\mathrm{Cay}(\Gamma, S)$ has vertex set
$\Gamma$ and edge set $\{\{g, gs\} : g \in \Gamma, s \in S\}$. It is
$|S|$-regular and connected iff $S$ generates $\Gamma$.
\end{definition}

\begin{example}[Familiar Cayley graphs]
\mbox{}
\begin{itemize}
  \item $C_n = \mathrm{Cay}(\mathbb Z_n, \{\pm 1\})$.
  \item Hypercube $Q_n = \mathrm{Cay}(\mathbb Z_2^n, \{e_1, \ldots, e_n\})$.
  \item Complete graph $K_n = \mathrm{Cay}(\Gamma, \Gamma \setminus \{e\})$
        for any group $\Gamma$ of order $n$.
  \item The \emph{circulant} graph $C_n(S)$ for $S \subseteq \{1, 2, \ldots, \lfloor n/2 \rfloor\}$
        is $\mathrm{Cay}(\mathbb Z_n, S \cup (-S))$.
\end{itemize}
\end{example}

\begin{theorem}[Cayley implies vertex-transitive]\label{thm:cayleyvt}
Every Cayley graph is vertex-transitive: left translation by any
$h \in \Gamma$ is an automorphism. The converse is false (e.g.\ the
Petersen graph is vertex-transitive but not Cayley for any group).
\end{theorem}

\subsubsection{Spectrum from Characters}

\begin{theorem}[Character formula for abelian Cayley spectra]\label{thm:cayleyabelian}
Let $\Gamma$ be a \emph{finite abelian} group and
$X = \mathrm{Cay}(\Gamma, S)$. The eigenvalues of $A(X)$ are exactly
\[
  \lambda_\chi \;=\; \sum_{s \in S} \chi(s),
  \qquad \chi \in \widehat{\Gamma},
\]
where $\widehat{\Gamma}$ is the group of characters of $\Gamma$. The
character $\chi$ itself, viewed as a vector $(\chi(g))_{g \in \Gamma}$,
is the eigenvector.
\end{theorem}

\begin{proof}
Since $\Gamma$ is abelian, characters form an orthonormal basis of
$\mathbb C^\Gamma$. Direct computation:
\[
  (A \chi)(g) = \sum_{s \in S} \chi(g s)
              = \chi(g) \sum_{s \in S} \chi(s)
              = \lambda_\chi \chi(g).
\]
\end{proof}

\begin{example}[Hypercube spectrum]
For $\Gamma = \mathbb Z_2^n$ the characters are
$\chi_T(x) = (-1)^{|T \cap x|}$ for $T \subseteq [n]$. The spectrum of
$Q_n$ is therefore
\[
  \lambda_T \;=\; \sum_{i = 1}^{n} \chi_T(e_i) \;=\; n - 2|T|,
  \qquad T \subseteq [n],
\]
giving eigenvalues $n - 2k$ ($0 \le k \le n$) with multiplicity
$\binom{n}{k}$.
\end{example}

\begin{example}[Cycle spectrum]
For $\Gamma = \mathbb Z_n$ the characters are $\chi_j(k) = \omega^{jk}$
with $\omega = e^{2\pi i/n}$. The spectrum of $C_n = \mathrm{Cay}(\mathbb Z_n, \{\pm 1\})$
is $\lambda_j = \omega^j + \omega^{-j} = 2\cos(2\pi j/n)$.
\end{example}

\subsubsection{Non-Abelian Cayley Graphs}

\begin{theorem}[Diaconis--Shahshahani trace formula]
Let $\Gamma$ be a finite (possibly non-abelian) group and
$S \subseteq \Gamma$ symmetric. Then the spectrum of
$\mathrm{Cay}(\Gamma, S)$ consists of the eigenvalues of the matrices
$\rho(\hat S) = \sum_{s \in S} \rho(s)$ as $\rho$ ranges over the
irreducible (unitary) representations of $\Gamma$, with multiplicities
$\dim(\rho)$.
\end{theorem}

\begin{example}[Random walks on $S_n$]
The transposition walk on $S_n$ corresponds to
$\mathrm{Cay}(S_n, S)$ with $S$ the set of all transpositions.
Diaconis and Shahshahani used the character formula to prove that the
mixing time is $\frac{1}{2} n \log n$, the canonical example of a
sharp \emph{cutoff phenomenon} for Markov chains.
\end{example}

\subsubsection{Vertex-Transitive Graphs and Cayley-ness}

\begin{theorem}[Sabidussi, 1958]
A vertex-transitive graph $X$ is a Cayley graph iff $\mathrm{Aut}(X)$
contains a subgroup acting regularly (transitively and freely) on
$V(X)$.
\end{theorem}

\begin{example}[Petersen is not Cayley]
$|\mathrm{Aut}(\text{Petersen})| = 120 = 5!$, with full automorphism
group $S_5$. There is no subgroup of $S_5$ of order $10$ acting
regularly on the $10$ vertices, so Petersen fails Sabidussi's
criterion.
\end{example}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Identify each of the following as a Cayley graph if possible: $K_4$,
$K_{3,3}$, $C_5$, $Q_3$, the Petersen graph.
\end{exercise}

\begin{exercise}
Compute the spectrum of the circulant graph $C_8(\{1, 3\})$ using
Theorem~\ref{thm:cayleyabelian}.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Show that $\mathrm{Cay}(\mathbb Z_n, S)$ is connected iff $\gcd(S) = 1$,
where $\gcd(S)$ is the GCD of the elements of $S \cup (-S) \cup \{n\}$.
\end{exercise}

\begin{exercise}
Verify that the Cayley graph $\mathrm{Cay}(D_8, \{r, r^{-1}, s\})$
(dihedral group with rotation $r$ of order $4$ and reflection $s$) is
isomorphic to the cube $Q_3$. Use Sabidussi's theorem to conclude that
$Q_3$ has at least two non-equivalent presentations as a Cayley graph.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove that for any abelian group $\Gamma$ and symmetric generating
set $S$, the second-largest eigenvalue $\lambda_2$ of $\mathrm{Cay}(\Gamma, S)$
satisfies
$\lambda_2 \ge |S| - 2 \pi^2 \cdot (|S|/\mathrm{diam}(X))^2 / 4$.
Conclude that abelian Cayley graphs cannot be expanders if the
generating set has bounded size.
\end{exercise}

\begin{exercise}
Use the character formula for non-abelian Cayley graphs to compute
the spectrum of the Cayley graph of $S_3$ with generating set
$\{(12), (13), (23)\}$. Identify the resulting graph.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw $\mathrm{Cay}(\mathbb Z_2^3, \{e_1, e_2, e_3\}) = Q_3$ and label
the eigenvectors $\chi_T$ for each $T \subseteq \{1, 2, 3\}$. Color
each vertex by the sign $\chi_T(g) = \pm 1$ for $T = \{1, 2\}$ and
verify that this colouring is balanced and orthogonal to the all-ones
vector.
\end{exercise}

\begin{exercise}
Draw the Cayley graph of $\mathbb Z_5$ with generating set
$\{\pm 1\}$ (the pentagon $C_5$). Mark the eigenvectors
$\chi_j(k) = e^{2\pi i j k/5}$ for $j = 1, 2$ and verify that the
corresponding eigenvalues are $2 \cos(2\pi/5)$ and $2 \cos(4\pi/5)$.
\end{exercise}

%====================================================================
\subsection{The Tutte Polynomial and Graphic Matroids}
%====================================================================

The Tutte polynomial $T_G(x, y)$ is the universal multiplicative
graph invariant on deletion--contraction. It specializes to the
chromatic polynomial, the flow polynomial, the all-terminal
reliability polynomial, the Jones polynomial of an alternating link,
and the Potts model partition function. Its natural setting is the
\emph{matroid} of $G$.

\subsubsection{Deletion, Contraction, and the Tutte Polynomial}

\begin{definition}[Deletion and contraction]
For a graph $G$ and edge $e \in E(G)$:
\begin{itemize}
  \item $G \setminus e$ is the graph obtained by removing $e$.
  \item $G / e$ is the graph obtained by contracting $e$ (identify
        its two endpoints, removing any loops created).
\end{itemize}
\end{definition}

\begin{definition}[Tutte polynomial]\label{def:tutte}
For a graph $G$ with vertex set $V$, edge set $E$, and rank
$r(G) = |V| - c(G)$ (where $c(G)$ is the number of connected
components), the \textbf{Tutte polynomial} is
\[
  T_G(x, y) \;=\; \sum_{A \subseteq E} (x - 1)^{r(E) - r(A)} (y - 1)^{|A| - r(A)},
\]
where $r(A) = |V| - c(V, A)$ is the rank of the spanning subgraph
$(V, A)$.
\end{definition}

\begin{theorem}[Deletion--contraction recursion]\label{thm:tuttedc}
For an edge $e \in E(G)$:
\begin{itemize}
  \item If $e$ is neither a loop nor a bridge:
        $T_G(x, y) = T_{G \setminus e}(x, y) + T_{G / e}(x, y)$.
  \item If $e$ is a bridge: $T_G(x, y) = x \cdot T_{G / e}(x, y)$.
  \item If $e$ is a loop: $T_G(x, y) = y \cdot T_{G \setminus e}(x, y)$.
  \item If $G$ has no edges: $T_G(x, y) = 1$.
\end{itemize}
These four rules determine $T_G$ uniquely.
\end{theorem}

\begin{example}[Tutte polynomial of small graphs]
\mbox{}
\begin{itemize}
  \item $T_{K_2}(x, y) = x$.
  \item $T_{K_3}(x, y) = x^2 + x + y$.
  \item $T_{K_4}(x, y) = x^3 + 3x^2 + 2x + 4xy + 2y + 3y^2 + y^3$.
\end{itemize}
\end{example}

\subsubsection{Specializations: Chromatic, Flow, Reliability}

\begin{theorem}[Whitney/Tutte specialization]\label{thm:tuttechrom}
For any graph $G$ on $n$ vertices,
\[
  \chi_G(k) \;=\; (-1)^{r(E)}\, k^{c(G)} \, T_G(1 - k, 0),
\]
where $\chi_G(k)$ is the chromatic polynomial counting proper
$k$-colorings.
\end{theorem}

\begin{theorem}[Flow polynomial]
The number of nowhere-zero $\mathbb Z_k$-flows on $G$ (with arbitrary
edge orientation) is
\[
  F_G(k) \;=\; (-1)^{|E| - r(E)}\, T_G(0, 1 - k).
\]
\end{theorem}

\begin{theorem}[Reliability polynomial]
If each edge of $G$ fails independently with probability $1 - p$, the
probability that the surviving subgraph remains connected (the
\textbf{all-terminal reliability}) is
\[
  R_G(p) \;=\; p^{r(E)} (1 - p)^{|E| - r(E)} \, T_G(1, 1/(1 - p)).
\]
\end{theorem}

\subsubsection{Whitney's Broken-Circuit Theorem}

\begin{definition}[Broken circuit]
Fix a total order on $E(G)$. A \textbf{broken circuit} is a set
$C \setminus \{e\}$ where $C$ is a cycle of $G$ and $e$ is the largest
edge of $C$.
\end{definition}

\begin{theorem}[Whitney's broken-circuit theorem]\label{thm:whitneybc}
For any total order on $E(G)$,
\[
  \chi_G(k) \;=\; \sum_{B \subseteq E,\ B \text{ contains no broken circuit}}
                  (-1)^{|B|}\, k^{n - |B|}.
\]
In particular, $\chi_G(k)$ has alternating-sign integer coefficients.
\end{theorem}

\begin{theorem}[Huh--Katz, 2012]
The coefficients of the chromatic polynomial form a \emph{log-concave}
sequence: if $\chi_G(k) = \sum_{i = 0}^{n} (-1)^i a_i k^{n - i}$, then
$a_i^2 \ge a_{i - 1} a_{i + 1}$ for $1 \le i \le n - 1$.
\end{theorem}

The Huh--Katz proof uses Hodge theory of the Chow ring of the
graphic matroid; this established the Rota--Heron--Welsh conjecture
in full generality.

\subsubsection{Graphic Matroids and Cycle/Bond Duality}

\begin{definition}[Matroid]\label{def:matroid}
A \textbf{matroid} on a finite set $E$ is a pair $(E, \mathcal I)$
where $\mathcal I \subseteq 2^E$ (the \textbf{independent sets})
satisfies:
\begin{itemize}
  \item $\emptyset \in \mathcal I$.
  \item If $I \in \mathcal I$ and $J \subseteq I$, then $J \in \mathcal I$.
  \item (\emph{Augmentation}) If $I, J \in \mathcal I$ with $|I| < |J|$,
        then there exists $e \in J \setminus I$ with $I \cup \{e\} \in \mathcal I$.
\end{itemize}
The maximal independent sets are the \textbf{bases}; the rank
function is $r(A) = \max\{|I| : I \subseteq A,\ I \in \mathcal I\}$.
\end{definition}

\begin{example}[Graphic matroid]
For a graph $G$, let $E = E(G)$ and
$\mathcal I = \{A \subseteq E : (V, A) \text{ is acyclic}\}$. This is
the \textbf{graphic matroid} $M(G)$. Bases of $M(G)$ are spanning
forests of $G$; the rank is $|V| - c(G)$.
\end{example}

\begin{theorem}[Cycle/bond duality on planar graphs]
For a planar graph $G$ with planar dual $G^*$,
$M(G^*) = M(G)^*$, where $M(G)^*$ is the matroid dual of $M(G)$.
Equivalently, cycles in $G$ correspond to bonds (minimal cuts) in
$G^*$ and vice versa.
\end{theorem}

\begin{theorem}[Tutte polynomial as matroid invariant]
The Tutte polynomial extends to all matroids via the same formula:
$T_M(x, y) = \sum_{A \subseteq E} (x - 1)^{r(E) - r(A)} (y - 1)^{|A| - r(A)}$.
For graphic matroids $M(G)$, this recovers $T_G(x, y)$. Matroid
duality interchanges $x$ and $y$:
$T_{M^*}(x, y) = T_M(y, x)$.
\end{theorem}

\subsubsection{Connections: Stanley's Chromatic Symmetric Function}

\begin{definition}[Chromatic symmetric function; Stanley, 1995]
For a graph $G$ on vertex set $V$, define the \textbf{chromatic
symmetric function}
\[
  X_G(x_1, x_2, \ldots) \;=\; \sum_{\kappa : V \to \mathbb Z_{\ge 1}\ \text{proper}}
                              \prod_{v \in V} x_{\kappa(v)}.
\]
\end{definition}

\begin{theorem}[Stanley]
$X_G$ is a symmetric function in $x_1, x_2, \ldots$, and $X_G(1, 1, \ldots, 1, 0, 0, \ldots) = \chi_G(k)$
when there are $k$ ones. Thus $X_G$ refines the chromatic polynomial.
\end{theorem}

\begin{conjecture}[Stanley's tree conjecture, 1995, OPEN]
If $T_1$ and $T_2$ are non-isomorphic trees on the same number of
vertices, then $X_{T_1} \ne X_{T_2}$.
\end{conjecture}

This conjecture has been verified by computer search for trees on up
to $\sim 30$ vertices but remains open. It would imply, in particular,
that the chromatic symmetric function distinguishes any pair of
non-isomorphic trees---a much finer invariant than the chromatic
polynomial $\chi_T(k) = k(k - 1)^{n - 1}$, which depends only on the
number of vertices.

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute $T_{C_4}(x, y)$ from Definition~\ref{def:tutte} by enumerating
all $2^4 = 16$ subsets of edges. Verify your answer against the
deletion--contraction recursion (Theorem~\ref{thm:tuttedc}).
\end{exercise}

\begin{exercise}
Use Theorem~\ref{thm:tuttechrom} to derive the chromatic polynomial
$\chi_{K_4}(k)$ from $T_{K_4}(x, y)$. Compare with the direct formula
$\chi_{K_n}(k) = k(k - 1)(k - 2) \cdots (k - n + 1)$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Prove the deletion--contraction recursion (Theorem~\ref{thm:tuttedc})
from Definition~\ref{def:tutte} by partitioning $A \subseteq E$ into
those containing $e$ and those not.
\end{exercise}

\begin{exercise}
Show that $T_G(2, 1) = $ number of spanning forests of $G$,
$T_G(1, 1) = $ number of spanning trees, and
$T_G(2, 0) = $ number of acyclic orientations of $G$ (Stanley).
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove Whitney's broken-circuit theorem (Theorem~\ref{thm:whitneybc})
by inclusion--exclusion on the cycles of $G$, ordered by their largest
edge.
\end{exercise}

\begin{exercise}
Show that the chromatic polynomial $\chi_G(k)$ has all real roots iff
$G$ is a chordal graph. (Hint: chromatic polynomials of chordal graphs
factor into linear pieces by perfect elimination ordering.)
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
For the planar graph $G = K_4$ minus an edge, draw both $G$ and its
planar dual $G^*$. Verify directly that $T_{G^*}(x, y) = T_G(y, x)$
by computing both polynomials and comparing.
\end{exercise}

\begin{exercise}
Draw the broken-circuit complex of the triangle $K_3$ with edges
ordered $e_1 < e_2 < e_3$. Identify the unique broken circuit
$\{e_1, e_2\}$, and verify
$\chi_{K_3}(k) = k^3 - 3k^2 + 2k$ from the broken-circuit sum of
Theorem~\ref{thm:whitneybc}.
\end{exercise}

%====================================================================
\subsection{Heaps of Pieces and Walk Generating Functions}
%====================================================================

This graduate-level section develops Viennot's theory of \emph{heaps
of pieces} and uses it to derive closed-form generating functions for
walks, matchings, and orthogonal polynomial moments on graphs. The
central theorems---Cartier--Foata's master inversion formula and
Flajolet's continued-fraction expansion---convert combinatorial counts
into rational and continued-fraction expressions, providing the
algebraic skeleton behind the matchings polynomial of \S 12.1 and the
Ihara zeta function of \S 12.7.

\subsubsection{Commutation Monoids and Heaps}

\begin{definition}[Commutation graph and partially commutative monoid]
\label{def:commgraph}
Let $A$ be a finite alphabet (the \emph{pieces}). A \textbf{commutation
graph} on $A$ is an undirected graph $C = (A, E_C)$. The
\textbf{partially commutative (Cartier--Foata) monoid} $M(C)$ is the
quotient of the free monoid $A^*$ by the relations
$ab \sim ba$ for every $\{a, b\} \in E_C$.
\end{definition}

\begin{definition}[Heap of pieces]
A \textbf{heap} on $C$ is an equivalence class of words in $M(C)$.
Pictorially: drop pieces one by one; two pieces $a, b$ rest at the
same height iff they commute (i.e.\ $\{a, b\} \in E_C$). A piece
\textbf{is at the bottom} of a heap $H$ if it has no piece below it.
A heap is \textbf{trivial} if all its pieces are at the bottom (i.e.\
pairwise commuting).
\end{definition}

\begin{example}[Heaps for monomer-dimer]
Let $G$ be a graph and let $A = E(G)$ with commutation graph
$C = $ the \emph{non-incidence graph} of $E(G)$ (edges $a, b$
commute iff they share no vertex). Then trivial heaps on $A$ are
exactly \textbf{matchings} of $G$. General heaps correspond to
sequences of edges placed with the constraint that overlapping edges
must respect a temporal order---i.e.\ \emph{lozenge tilings} of an
edge-disjoint pile.
\end{example}

\subsubsection{The Cartier--Foata Master Theorem}

\begin{theorem}[Cartier--Foata Master Theorem]\label{thm:cartierfoata}
Let $C$ be a commutation graph on $A$ with weights $w : A \to R$
(any commutative ring). Define
\[
  \mathcal H(z) = \sum_{H \text{ heap}} w(H)\, z^{|H|},
  \qquad
  \mathcal T(z) = \sum_{T \text{ trivial heap}} (-1)^{|T|} w(T)\, z^{|T|}.
\]
Then in the formal power series ring $R[[z]]$,
\[
  \mathcal H(z) \cdot \mathcal T(z) \;=\; 1,
  \qquad\text{equivalently}\qquad
  \mathcal H(z) \;=\; \frac{1}{\mathcal T(z)}.
\]
\end{theorem}

\begin{corollary}[Viennot's matching-polynomial formula]\label{cor:viennot}
For a graph $G$ with edge weights $w : E(G) \to R$, define the
\textbf{weighted matchings polynomial}
$\mu_w(G, z) = \sum_M (-1)^{|M|} \prod_{e \in M} w(e)\, z^{|M|}$,
the sum over all matchings $M \subseteq E(G)$. Then the generating
function of all heaps of edges (i.e.\ all walks-of-edges respecting
non-incidence) is exactly $1 / \mu_w(G, z)$.
\end{corollary}

\begin{remark}
Corollary~\ref{cor:viennot} explains the appearance of the matchings
polynomial $\mu(G, x)$ as a reciprocal: walks-on-edges satisfying
the heap commutation rule are dual to matchings in a precise
generating-function sense. This perspective recovers the
Heilmann--Lieb theorem (\ref{thm:hl}) without recourse to the
path-tree, and immediately generalises to $q$-analogues used in
mathematical physics.
\end{remark}

\subsubsection{Walks on Graphs as Heaps of Cycles}

\begin{definition}[Cycle heap]
For a graph $G$, the \textbf{cycle commutation graph} $C^{\mathrm{cyc}}(G)$
has vertex set $\mathcal C(G) = $ the set of (unoriented) simple
cycles of $G$. Two cycles commute (are connected in $C^{\mathrm{cyc}}$)
iff they are vertex-disjoint.
\end{definition}

\begin{theorem}[Sherman--Burgoyne; Viennot]\label{thm:cyclesheap}
The number of closed walks of length $k$ in $G$ equals
\[
  W_k(G) \;=\; [z^k]\, \frac{P(z)}{Q(z)},
\]
where $P(z)$ and $Q(z)$ are the generating functions of trivial heaps
of cycles, weighted respectively by $z^{\mathrm{length}}$ and
$(-1)^{\#\text{cycles}} z^{\mathrm{length}}$.
\end{theorem}

This formula expresses the trace $\mathrm{tr}(A^k)$ as a sum over
families of vertex-disjoint cycles---a graph-theoretic version of the
Lindstr\"om--Gessel--Viennot lemma.

\subsubsection{Flajolet's Continued-Fraction Theorem}

\begin{theorem}[Flajolet, 1980]\label{thm:flajolet}
Let $G$ be the path graph $\{0, 1, 2, \ldots\}$ with weighted self-loops
$b_n$ at vertex $n$ and weighted edges $\lambda_n$ between $n - 1$ and
$n$. The generating function of weighted closed walks at $0$ is the
Jacobi continued fraction
\[
  \sum_{k \ge 0} m_k z^k \;=\; \cfrac{1}{1 - b_0 z - \cfrac{\lambda_1 z^2}{1 - b_1 z - \cfrac{\lambda_2 z^2}{1 - b_2 z - \ddots}}},
\]
where $m_k$ is the total weight of all closed walks at $0$ of length
$k$. Equivalently, $(m_k)$ is the moment sequence of the orthogonal
polynomials with three-term recurrence
$P_{n+1}(x) = (x - b_n) P_n(x) - \lambda_n P_{n-1}(x)$.
\end{theorem}

\begin{corollary}[Closed walks $\leftrightarrow$ orthogonal polynomial moments]
Closed walks on weighted half-line graphs are in bijection with
\textbf{Motzkin paths}, whose moments coincide with classical
orthogonal-polynomial moments (Hermite, Laguerre, Meixner, etc.).
This is the algebraic root of why the matching polynomial of $K_n$ is
Hermite, of $P_n$ is Chebyshev, of $K_{m,n}$ is Laguerre.
\end{corollary}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
List all heaps of pieces on the path commutation graph $C = P_3$ with
pieces $\{a, b, c\}$ and weights $w(a) = w(b) = w(c) = 1$, of total
size $\le 3$. Verify the trivial-heap generating function
$\mathcal T(z) = 1 - 3z + z^2$ and the master theorem
$\mathcal H(z) = 1/(1 - 3z + z^2)$.
\end{exercise}

\begin{exercise}
For the matching polynomial of $K_3$, verify that
$1/\mu(K_3, z) = 1/(1 - 3z^2)$ in formal power series, and interpret
the coefficients as counts of weighted heaps of edges in $K_3$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Use Theorem~\ref{thm:cartierfoata} to derive the Lindstr\"om--Gessel--Viennot
lemma in the special case of vertex-disjoint paths in a planar
acyclic digraph.
\end{exercise}

\begin{exercise}
Apply Flajolet's theorem (\ref{thm:flajolet}) with $b_n = 0$ and
$\lambda_n = n$ to recover the moment generating function of the
Hermite polynomials, $\sum_k (2k - 1)!! \, z^{2k}$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove Theorem~\ref{thm:cyclesheap} by induction on the trivial-heap
size, using Theorem~\ref{thm:cartierfoata} on the cycle commutation
graph $C^{\mathrm{cyc}}(G)$.
\end{exercise}

\begin{exercise}
Generalise Flajolet's theorem to closed walks on a weighted bipartite
graph: derive the Stieltjes continued fraction
$\sum m_k z^k = 1/(1 - c_1 z/(1 - c_2 z/(1 - \cdots)))$ and identify
the corresponding orthogonal polynomials.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw a sample heap of $4$ pieces on the commutation graph
$C = $ disjoint union of $K_2$ and an isolated vertex (i.e.\ pieces
$a, b, c$ with $a \sim b$ and $c$ isolated). Sketch the heap as a
``stack'' picture and identify which pieces are at the bottom.
\end{exercise}

\begin{exercise}
Sketch the cycle commutation graph $C^{\mathrm{cyc}}(K_4)$. Identify
the simple cycles of $K_4$ (four $C_3$'s, three $C_4$'s) and the
disjointness relations between them. Verify Theorem~\ref{thm:cyclesheap}
gives $W_4(K_4) = $ number of closed walks of length $4$ in $K_4$
($= \mathrm{tr}(A(K_4)^4) = 3^4 + (-1)^4 + (-1)^4 + (-1)^4 = 84$).
\end{exercise}

%====================================================================
\subsection{The Ihara Zeta Function of a Graph}
%====================================================================

The Ihara zeta function is the graph-theoretic analogue of the
Riemann zeta function. Originally introduced by Ihara (1966) in the
context of $p$-adic groups and rediscovered combinatorially by
Sunada, Hashimoto, Bass, and Stark--Terras, it packages \emph{prime
non-backtracking closed walks} into a generating function whose
poles encode spectral properties of the graph. The Ramanujan property
of \S 11.4 turns out to be exactly the ``Riemann hypothesis'' for the
Ihara zeta.

\subsubsection{Non-Backtracking Walks and Primes}

\begin{definition}[Non-backtracking walk; prime walk]
A walk $w = v_0 e_1 v_1 e_2 v_2 \cdots e_k v_k$ in $G$ is
\textbf{non-backtracking} if $e_{i+1} \ne \overline{e_i}$ for every
$i$, where $\overline{e}$ denotes the reversed direction of edge $e$.
A non-backtracking closed walk is \textbf{prime} (or \emph{primitive})
if it is not a proper power of a shorter closed walk and we identify
walks differing only by cyclic rotation. Let $\mathcal P(G)$ denote
the set of equivalence classes of prime closed walks.
\end{definition}

\begin{definition}[Ihara zeta function]\label{def:ihara}
For a finite connected graph $G$ with no degree-$1$ vertex, the
\textbf{Ihara zeta function} is
\[
  Z_G(u) \;=\; \prod_{[C] \in \mathcal P(G)} \big(1 - u^{|C|}\big)^{-1}.
\]
\end{definition}

\begin{example}[Cycle graph]
For $C_n$ ($n \ge 3$), there are exactly two prime closed walks (the
clockwise and counterclockwise traversals), each of length $n$. Hence
$Z_{C_n}(u) = (1 - u^n)^{-2}$.
\end{example}

\subsubsection{The Determinantal Formula}

\begin{theorem}[Ihara--Bass formula]\label{thm:iharabass}
Let $G$ be a finite connected graph with $n$ vertices, $m$ edges, and
adjacency matrix $A$. Set $r = m - n + 1$ (the cycle rank), and let
$D$ be the degree matrix. Then
\[
  Z_G(u)^{-1} \;=\; (1 - u^2)^{r - 1} \cdot \det\!\big(I - u A + u^2 (D - I)\big).
\]
For a $d$-regular graph, the formula simplifies to
\[
  Z_G(u)^{-1} \;=\; (1 - u^2)^{r - 1} \cdot \det\!\big(I - u A + u^2 (d - 1) I\big).
\]
\end{theorem}

\begin{proof}[Proof sketch via the Hashimoto matrix]
Define the \textbf{non-backtracking matrix} $B \in \{0, 1\}^{2m \times 2m}$
indexed by directed edges:
$B_{(u,v),(v,w)} = 1$ iff $w \ne u$.
The set of non-backtracking closed walks of length $k$ is exactly
$\mathrm{tr}(B^k)$, so by the standard zeta-function identity
$Z_G(u) = 1/\det(I - u B)$. Bass's three-term identity then relates
$\det(I - uB)$ to the smaller $n \times n$ determinant of
Theorem~\ref{thm:iharabass}.
\end{proof}

\subsubsection{Riemann Hypothesis for Graphs}

\begin{definition}[Graph Riemann hypothesis]
A $d$-regular graph $G$ \textbf{satisfies the Riemann Hypothesis}
(GRH) if all non-trivial poles of $Z_G(u)$ lie on the circle
$|u| = 1/\sqrt{d - 1}$.
\end{definition}

\begin{theorem}[Ramanujan $\Leftrightarrow$ GRH]\label{thm:ramanujangrh}
A connected $d$-regular graph $G$ is Ramanujan
(Definition~\ref{def:ramanujan} -- see \S 11.4) if and only if it
satisfies the graph Riemann Hypothesis. Equivalently:
\[
  G \text{ Ramanujan} \;\Longleftrightarrow\;
  \text{spec}(A) \subseteq \{d\} \cup [-2\sqrt{d - 1},\, 2\sqrt{d - 1}].
\]
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:iharabass}, the poles of $Z_G(u)$ for a
$d$-regular graph are the reciprocals of the eigenvalues of $A$
\emph{transformed} by the substitution
$\lambda \mapsto u$ such that $\lambda = (1 + u^2(d - 1))/u$. Solving
this quadratic, $|u| = 1/\sqrt{d-1}$ corresponds exactly to
$|\lambda| \le 2\sqrt{d - 1}$.
\end{proof}

\begin{remark}[Non-backtracking spectrum and community detection]
Krzakala et al.\ (2013) showed that the spectrum of the Hashimoto
matrix $B$ (rather than the adjacency matrix $A$) detects communities
in sparse random graphs all the way down to the Kesten--Stigum
threshold, which is below the threshold detectable by $A$ alone. The
non-backtracking matrix has thus become the central tool in modern
spectral methods for community detection in stochastic block models.
\end{remark}

\subsubsection{Functional Equation and Special Values}

\begin{theorem}[Functional equation]
For a $d$-regular graph,
\[
  Z_G(u) \cdot Z_G\!\left(\frac{1}{(d - 1) u}\right) = (\text{simple factor})
\]
analogous to the symmetry $\zeta(s) \leftrightarrow \zeta(1 - s)$ for
Riemann zeta. The exact form involves $(1 - u^2)$, $((d-1)u^2 - 1)$,
and powers depending on $r$.
\end{theorem}

\begin{theorem}[Special value $u = 1$]
For a connected graph $G$,
\[
  Z_G(1)^{-1} \;=\; (-2)^r \cdot \tau(G),
\]
where $\tau(G)$ is the number of spanning trees. This is a graph-zeta
analogue of the class number formula.
\end{theorem}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute $Z_{K_4}(u)$ explicitly using the Ihara--Bass formula. ($K_4$
is $3$-regular with adjacency spectrum $\{3, -1, -1, -1\}$ and
$r = 6 - 4 + 1 = 3$.)
\end{exercise}

\begin{exercise}
Verify Theorem~\ref{thm:ramanujangrh} for the Petersen graph: its
spectrum $\{3, 1^{(5)}, -2^{(4)}\}$ lies in
$\{3\} \cup [-2\sqrt{2}, 2\sqrt{2}]$, so it is Ramanujan, and hence
all non-trivial poles of $Z_P(u)$ lie on $|u| = 1/\sqrt{2}$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Prove the cycle formula $Z_{C_n}(u) = (1 - u^n)^{-2}$ directly from
Definition~\ref{def:ihara} by enumerating prime closed walks.
\end{exercise}

\begin{exercise}
Verify Bass's three-term identity for the Hashimoto matrix:
$\det(I - u B) = (1 - u^2)^{r - 1} \det(I - u A + u^2 (D - I))$, by
explicitly computing both sides for $K_3$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Use the non-backtracking matrix $B$ and the Perron--Frobenius theorem
to prove that for a connected $d$-regular non-bipartite graph, the
spectral radius of $B$ is exactly $d - 1$, and the corresponding
eigenvector is the all-ones vector.
\end{exercise}

\begin{exercise}
Show that the Ihara zeta function of an infinite $d$-regular tree
has no poles, in agreement with the universal-cover spectrum
$[-2\sqrt{d - 1}, 2\sqrt{d - 1}]$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
For the graph $K_4$, list all prime closed walks of length $\le 6$
(up to cyclic rotation and inversion), grouped by length. Compute the
first few coefficients of $\log Z_{K_4}(u) = \sum_k N_k u^k / k$
where $N_k$ counts non-backtracking closed walks of length $k$.
\end{exercise}

\begin{exercise}
Sketch the Hashimoto matrix $B$ for the triangle $K_3$, showing the
$6 \times 6$ structure indexed by the $6$ directed edges. Verify by
hand that $\det(I - uB) = (1 - u^2)(1 - u)^2(1 - 2u)(1 + u + 2u^2)$
or whatever the correct factorisation is for $K_3$.
\end{exercise}

%====================================================================
\subsection{Combinatorial Hopf Algebras of Graphs}
%====================================================================

A \emph{combinatorial Hopf algebra} packages a family of combinatorial
objects together with operations of \emph{merging} (product) and
\emph{splitting} (coproduct) compatible with one another. The graph
Hopf algebra of Schmitt (1994) and the chromatic Hopf algebra of
Stanley (1995) put the chromatic polynomial and chromatic symmetric
function into a clean algebraic frame---and the
Aguiar--Bergeron--Sottile theorem (2006) shows that this frame is
\emph{universal}: every combinatorial Hopf algebra has a canonical
character, and the chromatic symmetric function is the image of the
graph Hopf algebra under it.

\subsubsection{Hopf Algebras: A Five-Line Reminder}

\begin{definition}[Hopf algebra]\label{def:hopf}
A \textbf{Hopf algebra} over a field $k$ is a $k$-vector space $H$
equipped with linear maps:
\begin{itemize}
  \item product $\mu : H \otimes H \to H$ (associative),
  \item unit $\eta : k \to H$,
  \item coproduct $\Delta : H \to H \otimes H$ (coassociative),
  \item counit $\varepsilon : H \to k$,
  \item antipode $S : H \to H$,
\end{itemize}
satisfying the standard compatibility axioms (the antipode
satisfies $\mu(S \otimes \mathrm{id})\Delta = \mu(\mathrm{id} \otimes S)\Delta = \eta \varepsilon$).
A Hopf algebra is \textbf{graded} if $H = \bigoplus_{n \ge 0} H_n$
with all structure maps respecting the grading and $H_0 = k$, and is
\textbf{connected} if $H_0 = k \cdot 1$.
\end{definition}

\begin{example}[Symmetric functions]
The ring $\Lambda$ of symmetric functions is a graded connected
Hopf algebra: the coproduct is $\Delta(p_n) = p_n \otimes 1 + 1 \otimes p_n$
on the power-sum basis (i.e.\ the $p_n$ are \emph{primitive}). It is
self-dual: $\Lambda \cong \Lambda^*$ as Hopf algebras, with the Hall
inner product realising the duality.
\end{example}

\subsubsection{The Graph Hopf Algebra of Schmitt}

\begin{definition}[Schmitt's graph Hopf algebra]\label{def:schmitt}
Let $\mathcal G_n$ be the set of (isomorphism classes of) finite
simple graphs with vertex set $[n]$, and let
$\mathcal G = \bigoplus_{n \ge 0} k\, \mathcal G_n$. Define:
\begin{itemize}
  \item Product: $G \cdot H = G \sqcup H$ (disjoint union, relabelled).
  \item Coproduct:
        $\Delta(G) = \sum_{S \subseteq V(G)} G[S] \otimes G[V \setminus S]$,
        the sum over all bipartitions of $V(G)$ into induced subgraphs.
  \item Unit: the empty graph; counit: $\varepsilon(G) = [G = \emptyset]$.
\end{itemize}
Then $(\mathcal G, \mu, \Delta, \eta, \varepsilon)$ is a graded
connected commutative cocommutative Hopf algebra.
\end{definition}

\begin{theorem}[Antipode formula; Schmitt, Humpert--Martin]
The antipode of $\mathcal G$ admits a cancellation-free formula:
\[
  S(G) \;=\; \sum_{F : \mathrm{flat}} (-1)^{c(F)} \, a(F) \, G/F,
\]
where the sum is over flats $F$ of the cycle matroid of $G$, $c(F)$
is the number of bonds of $F$, $a(F)$ is the number of acyclic
orientations of the contraction $G/F$, and $G/F$ is the quotient
graph.
\end{theorem}

\subsubsection{Characters and the Aguiar--Bergeron--Sottile Theorem}

\begin{definition}[Character]
A \textbf{character} on a graded connected Hopf algebra $H$ is a
multiplicative linear functional $\zeta : H \to k$ with
$\zeta(1) = 1$ and $\zeta(xy) = \zeta(x) \zeta(y)$.
\end{definition}

\begin{theorem}[Aguiar--Bergeron--Sottile, 2006]\label{thm:abs}
The Hopf algebra $\mathrm{QSym}$ of quasisymmetric functions, equipped
with the character $\zeta_Q : F_\alpha \mapsto [\alpha = (n)]$ (the
``forget compositions'' character), is the \textbf{terminal object}
in the category of combinatorial Hopf algebras: for every
$(H, \zeta_H)$, there is a unique Hopf algebra map $\Psi : H \to \mathrm{QSym}$
making
\[
  \zeta_Q \circ \Psi \;=\; \zeta_H.
\]
\end{theorem}

\subsubsection{Application: Stanley's Chromatic Symmetric Function}

\begin{definition}[Chromatic character on $\mathcal G$]
Define $\zeta_{\mathrm{chr}} : \mathcal G \to \mathbb Q$ by
$\zeta_{\mathrm{chr}}(G) = $ number of proper $1$-colorings of $G$
(which is $1$ if $G$ has no edges and $0$ otherwise). Then
$\zeta_{\mathrm{chr}}$ is a character on $\mathcal G$.
\end{definition}

\begin{theorem}[Stanley, 1995; ABS reformulation]\label{thm:stanleychr}
The unique Hopf algebra morphism $\Psi : \mathcal G \to \mathrm{QSym}$
satisfying $\zeta_Q \circ \Psi = \zeta_{\mathrm{chr}}$ sends $G$ to the
\textbf{chromatic symmetric function}:
\[
  \Psi(G) \;=\; X_G(x_1, x_2, \ldots) \;=\; \sum_{\kappa \text{ proper}} \prod_{v \in V(G)} x_{\kappa(v)}.
\]
In particular $X_G$ is symmetric (not merely quasisymmetric).
\end{theorem}

\begin{remark}[Schur positivity is open]
Stanley conjectured (and verified for small $n$) that $X_G$ is a
non-negative integer combination of Schur functions $s_\lambda$ for
broad classes of graphs (e.g.\ \emph{claw-free} graphs and
\emph{$(3+1)$-free} graphs). This is the celebrated
\textbf{Stanley--Stembridge conjecture}, recently proved by
Hikita (2024).
\end{remark}

\subsubsection{Other Combinatorial Hopf Algebras}

\begin{example}[Posets]
The Hopf algebra of finite posets, with disjoint union as product and
$\Delta(P) = \sum_{I \text{ order ideal}} I \otimes (P \setminus I)$,
maps under ABS to the $P$-partition generating function (Stanley,
Gessel).
\end{example}

\begin{example}[Matroids]
Schmitt's matroid Hopf algebra has $\Delta(M) = \sum_{A \subseteq E(M)} M|_A \otimes M/A$
and maps to QSym via the rank generating function. The image recovers
the Tutte polynomial as a quasisymmetric specialisation.
\end{example}

\begin{example}[The Connes--Kreimer Hopf algebra of trees]
Used in renormalisation theory of quantum field theory. Coproduct
extracts admissible cuts of a rooted tree; antipode encodes the
``BPHZ'' renormalisation procedure.
\end{example}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Verify that the coproduct
$\Delta(G) = \sum_{S \subseteq V} G[S] \otimes G[V \setminus S]$ is
coassociative on the path $P_3$.
\end{exercise}

\begin{exercise}
Compute $\Delta(K_3)$ in $\mathcal G$ explicitly. The result should
be a sum of $2^3 = 8$ terms, including the trivial pieces
$K_3 \otimes \emptyset$ and $\emptyset \otimes K_3$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Verify Theorem~\ref{thm:stanleychr} for $G = K_3$: compute $X_{K_3}$
directly from the definition, then show that the unique character-preserving
map $\mathcal G \to \mathrm{QSym}$ produces the same symmetric function.
\end{exercise}

\begin{exercise}
Show that the chromatic polynomial $\chi_G(k)$ is the principal
specialisation of $X_G$:
$\chi_G(k) = X_G(1, 1, \ldots, 1, 0, 0, \ldots)$ with $k$ ones.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove the antipode formula for $\mathcal G$ (Schmitt's formula): for
any graph $G$,
$S(G) = \sum_{F} (-1)^{|E(F)|} \,
\mu(\hat 0, F)\, G/F$
where the sum is over flats of the bond lattice and $\mu$ is the
M\"obius function.
\end{exercise}

\begin{exercise}
Use the Aguiar--Bergeron--Sottile universal property
(Theorem~\ref{thm:abs}) to derive the Stanley--Reisner $h$-vector
formula for the order complex of a poset.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw all $8$ pairs of induced subgraphs of $C_4 = $ the $4$-cycle that
appear in $\Delta(C_4)$. Group them by the size of the left factor and
verify the $\Delta$ sum.
\end{exercise}

\begin{exercise}
Sketch the Hopf algebra structure on rooted trees of size $\le 3$ via
the Connes--Kreimer coproduct $\Delta(T) = \sum_c T_c^{\mathrm{trunk}} \otimes T_c^{\mathrm{branches}}$
over admissible cuts $c$. Compute $\Delta$ on the path $P_3$ and the
star $S_3$.
\end{exercise}

%====================================================================
\subsection{Hodge Theory of Matroids and Log-Concavity}
%====================================================================

The chromatic polynomial of a graph---and more generally the
characteristic polynomial of a matroid---has integer coefficients with
\emph{alternating signs} (Whitney, 1932). Stanley conjectured in 1980
that these coefficients in absolute value form a \emph{log-concave}
sequence, and Mason conjectured the same for the $f$-vector of any
matroid. Both conjectures are now theorems, proved using
combinatorial Hodge theory by Adiprasito--Huh--Katz (2018), bringing
algebraic-geometric machinery to bear on a purely combinatorial
question. This section is the most advanced in the chapter and
sketches the key ideas at a graduate level.

\subsubsection{Log-Concavity Conjectures}

\begin{definition}[Log-concave sequence]
A finite sequence $a_0, a_1, \ldots, a_n$ of non-negative reals is
\textbf{log-concave} if $a_i^2 \ge a_{i-1} a_{i+1}$ for $1 \le i \le n - 1$.
It is \textbf{ultra log-concave} if $a_i / \binom{n}{i}$ is
log-concave.
\end{definition}

\begin{theorem}[Heron--Rota--Welsh / Adiprasito--Huh--Katz, 2018]\label{thm:ahk}
Let $M$ be a matroid of rank $r$ on ground set $E$, and let
$\chi_M(t) = \sum_{i = 0}^{r} (-1)^i w_i \, t^{r - i}$ be its
characteristic polynomial, where $w_i \ge 0$ are the
\textbf{Whitney numbers of the first kind}. Then $(w_0, w_1, \ldots, w_r)$
is a log-concave sequence with no internal zeros.
\end{theorem}

\begin{corollary}[Stanley's chromatic log-concavity, 1980]
For any graph $G$ on $n$ vertices, the coefficients of the chromatic
polynomial $\chi_G(k) = \sum_{i = 0}^{n - 1} (-1)^i a_i k^{n - i}$
satisfy $a_i^2 \ge a_{i - 1} a_{i + 1}$.
\end{corollary}

\begin{corollary}[Mason's conjecture; Adiprasito--Huh--Katz; Anari--Liu--Oveis Gharan--Vinzant; Br\"and\'en--Huh, all 2018]
For any matroid $M$, the $f$-vector $(f_0, f_1, \ldots, f_r)$
counting independent sets by size is ultra-log-concave.
\end{corollary}

\subsubsection{The Chow Ring of a Matroid}

\begin{definition}[Bergman fan and Chow ring; Feichtner--Yuzvinsky, 2004]\label{def:chowring}
Let $M$ be a loopless matroid on $E = [n]$ with lattice of flats $L_M$
and rank function $r$. The \textbf{Chow ring} $A^*(M)$ is the
$\mathbb Z$-graded commutative ring
\[
  A^*(M) \;=\; \mathbb Z[x_F : F \in L_M, F \ne \hat 0, F \ne \hat 1]\, /\, (I_1 + I_2),
\]
where:
\begin{itemize}
  \item $I_1$ is generated by $x_F x_{F'}$ for incomparable flats
        $F, F'$ (the \textbf{Stanley--Reisner relations}).
  \item $I_2$ is generated by
        $\sum_{F \ni i} x_F - \sum_{F \ni j} x_F$ for $i, j \in E$
        (the \textbf{linear relations}).
\end{itemize}
The grading places $\deg(x_F) = 1$. The ring is concentrated in
degrees $0, 1, \ldots, r - 1$, and
$\dim_{\mathbb Q} A^k(M) \otimes \mathbb Q$ is the (signed) coefficient
$|w_k|$ of the characteristic polynomial.
\end{definition}

\begin{theorem}[Adiprasito--Huh--Katz Hodge package]\label{thm:hodgepkg}
For every loopless matroid $M$ of rank $r$, the Chow ring $A^*(M)_{\mathbb R}$
satisfies the analogue of the K\"ahler package from algebraic
geometry:
\begin{enumerate}
  \item \textbf{Poincar\'e duality}: there is a degree map
        $\mathrm{deg} : A^{r - 1}(M) \to \mathbb R$ such that the
        induced pairing $A^k \otimes A^{r - 1 - k} \to \mathbb R$ is
        non-degenerate.
  \item \textbf{Hard Lefschetz}: for any \emph{strictly convex}
        element $\ell \in A^1(M)$, multiplication by $\ell^{r - 1 - 2k}$
        is an isomorphism $A^k \to A^{r - 1 - k}$.
  \item \textbf{Hodge--Riemann relations}: the bilinear form
        $(\eta_1, \eta_2) \mapsto (-1)^k \mathrm{deg}(\eta_1 \cdot \ell^{r - 1 - 2k} \cdot \eta_2)$
        is positive definite on the primitive part
        $\ker(\ell^{r - 2k} \cdot - : A^k \to A^{r - k})$.
\end{enumerate}
\end{theorem}

\subsubsection{From Hodge--Riemann to Log-Concavity}

\begin{theorem}[How HR implies log-concavity]\label{thm:hrlc}
Let $A = \bigoplus A^k$ be a graded ring satisfying the Hodge--Riemann
relations with respect to $\ell \in A^1$, and let $\alpha, \beta \in A^1$
have non-negative pairings with $\ell$. Then for any $0 \le k \le r - 2$,
\[
  \mathrm{deg}(\alpha \beta \ell^{r - 2 - k}) \cdot \mathrm{deg}(\alpha \beta \ell^{r - 2 - k})
  \;\ge\; \mathrm{deg}(\alpha^2 \ell^{r - 2 - k}) \cdot \mathrm{deg}(\beta^2 \ell^{r - 2 - k}).
\]
This Cauchy--Schwarz-type inequality is precisely the log-concavity
relation for the corresponding combinatorial sequence.
\end{theorem}

\begin{proof}[Proof sketch of Theorem~\ref{thm:hrlc}]
The Hodge--Riemann relations imply that the bilinear form
$Q(\eta_1, \eta_2) = \mathrm{deg}(\eta_1 \cdot \eta_2 \cdot \ell^{r - 2 - k})$
on $A^1$ has signature $(1, \dim A^1 - 1)$. The reverse Cauchy--Schwarz
for forms of this signature gives the log-concavity statement.
\end{proof}

\begin{remark}[The Hodge--Riemann elephant in the room]
For \emph{realizable} matroids (those arising from a hyperplane
arrangement over $\mathbb C$), the Chow ring $A^*(M)$ coincides with
the cohomology ring of a projective variety (the \emph{wonderful
compactification} of the arrangement complement, due to De
Concini--Procesi). For such varieties Hodge--Riemann is a classical
theorem of K\"ahler geometry.
The miracle of Adiprasito--Huh--Katz is that the K\"ahler package
holds combinatorially, even when the matroid is \emph{non-realizable}
(e.g.\ the V\'amos matroid, the non-Pappus matroid). The proof uses
an intricate inductive flip argument on the lattice of flats.
\end{remark}

\subsubsection{Spectral Independence: A Modern Cousin}

\begin{theorem}[Anari--Liu--Oveis Gharan--Vinzant, 2019]\label{thm:specind}
Let $M$ be a matroid of rank $r$ on $n$ elements and let $\mu$ be the
uniform distribution on bases of $M$. The down-up walk on bases of
$M$ has spectral gap $\Omega(1/r)$ and mixes in time
$O(r \log n)$.
\end{theorem}

\begin{remark}
Theorem~\ref{thm:specind} resolved a conjecture of Mihail--Vazirani
(1989) on rapid mixing for the bases-exchange walk. The proof uses
the matroid \emph{log-concavity} from Theorem~\ref{thm:ahk} together
with a new technique called \textbf{spectral independence}, which has
since become foundational in the modern probabilistic method (e.g.\
for sampling proper colorings, independent sets, and Ising / Potts
models).
\end{remark}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Verify Stanley's log-concavity conjecture by hand for the chromatic
polynomial of $C_4$:
$\chi_{C_4}(k) = (k - 1)^4 + (k - 1) = k^4 - 4k^3 + 6k^2 - 4k + (k - 1) + 1$.
Identify the absolute coefficients and check $a_i^2 \ge a_{i-1} a_{i+1}$.
\end{exercise}

\begin{exercise}
For the uniform matroid $U_{3,5}$ of rank $3$ on $5$ elements,
compute the $f$-vector $(f_0, f_1, f_2, f_3) = (1, 5, 10, 10)$ and
verify ultra-log-concavity: $f_i / \binom{5}{i}$ should be log-concave.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Prove that log-concavity is preserved under \emph{Hadamard products}:
if $(a_i)$ and $(b_i)$ are non-negative log-concave, so is
$(a_i b_i)$.
\end{exercise}

\begin{exercise}
Use Theorem~\ref{thm:hrlc} (Hodge--Riemann implies log-concavity) to
re-derive the log-concavity of the coefficients of the chromatic
polynomial of a cycle $C_n$ via the Chow ring of the cycle matroid
$U_{n - 1, n}$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Construct the Chow ring $A^*(U_{2, 3})$ of the uniform matroid of
rank $2$ on $3$ elements explicitly, including its Stanley--Reisner
relations. Verify Poincar\'e duality and Hard Lefschetz directly.
\end{exercise}

\begin{exercise}
Show that the matroid Hodge--Riemann relations of
Theorem~\ref{thm:hodgepkg} reduce, in the case of a realisable
matroid, to the Hodge--Riemann relations on the cohomology of the
De Concini--Procesi wonderful compactification.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the lattice of flats of the cycle matroid of $K_4$ (which is
the Boolean lattice $B_3$). Identify the Whitney numbers
$w_i = $ number of flats of rank $i$, and verify
$(1, 7, 6, 1)$ is log-concave.
\end{exercise}

\begin{exercise}
Sketch the geometric picture: for the rank-$2$ uniform matroid
$U_{2, 4}$, draw the Bergman fan in $\mathbb R^3$ as a tropical
linear space, and identify the cones corresponding to the flats of
$U_{2, 4}$.
\end{exercise}

%====================================================================
\subsection{The Permanent, Matching Counting, and \#P-Hardness}
%====================================================================

\paragraph{Graph-theoretic motivation.}
The number of perfect matchings of a bipartite graph equals the
\emph{permanent} of its biadjacency matrix---an algebraic invariant
that looks like the determinant but lacks its sign cancellations.
This single observation places the seemingly innocuous problem
``count the perfect matchings of $G$'' at the head of an entire
complexity-theoretic hierarchy: it is the canonical \#P-complete
problem (Valiant, 1979), in stark contrast to ``decide whether $G$
has a perfect matching'' (which is in $\mathsf P$ by the Hopcroft--Karp
algorithm of \S 7).

\subsubsection{The Permanent and Bipartite Matchings}

\begin{definition}[Permanent]\label{def:permanent}
For an $n \times n$ matrix $A = (a_{ij})$ over a commutative ring,
the \textbf{permanent} is
\[
  \mathrm{perm}(A) \;=\; \sum_{\sigma \in S_n} \prod_{i = 1}^{n} a_{i, \sigma(i)}.
\]
This is the determinant formula with the sign $\mathrm{sgn}(\sigma)$
removed.
\end{definition}

\begin{theorem}[Permanent counts perfect matchings]\label{thm:permmatch}
Let $G = (A \cup B, E)$ be a bipartite graph with $|A| = |B| = n$ and
biadjacency matrix $M \in \{0, 1\}^{n \times n}$ (with $M_{ij} = 1$
iff $a_i b_j \in E$). Then
\[
  \mathrm{perm}(M) \;=\; \#\{\text{perfect matchings of } G\}.
\]
\end{theorem}

\begin{proof}
A non-zero contribution from $\sigma \in S_n$ requires
$M_{i, \sigma(i)} = 1$ for all $i$, i.e.\ $\{a_i b_{\sigma(i)}\}$ is
a perfect matching. Conversely, every perfect matching is encoded by
exactly one permutation. Each contributes $1$.
\end{proof}

\subsubsection{Ryser's Inclusion--Exclusion Formula}

\begin{theorem}[Ryser, 1963]\label{thm:ryser}
For an $n \times n$ matrix $A$,
\[
  \mathrm{perm}(A) \;=\; (-1)^n \sum_{S \subseteq [n]} (-1)^{|S|} \prod_{i = 1}^{n} \Big(\sum_{j \in S} a_{ij}\Big).
\]
This formula computes the permanent in $O(n \cdot 2^n)$ arithmetic
operations---the fastest known algorithm for general matrices.
\end{theorem}

\begin{proof}[Proof sketch]
By inclusion--exclusion on the columns. Subtract off contributions
from $\sigma$ avoiding column $j$, then add back those avoiding two
columns, and so on.
\end{proof}

\begin{remark}
Despite its exponential complexity, Ryser's formula is the basis of
many practical permanent codes; the Glynn--Kogan formula gives a
similar bound with parity-check refinement.
\end{remark}

\subsubsection{Valiant's Theorem: \#P-Completeness}

\begin{theorem}[Valiant, 1979]\label{thm:valiant}
Computing the permanent of a $0/1$ matrix is \textbf{\#P-complete},
even when restricted to matrices arising from $3$-regular bipartite
graphs.
\end{theorem}

\begin{remark}
\#P is the counting analogue of $\mathsf{NP}$: a function $f$ is in
\#P if there is a polynomial-time non-deterministic machine $M$ such
that $f(x) = $ number of accepting paths of $M$ on $x$. Valiant's
theorem says that the seemingly innocent problem ``count perfect
matchings'' is \textbf{strictly harder than} all of $\mathsf{NP}$
unless the polynomial hierarchy collapses. In particular,
$\mathsf{P}^{\#P}$ contains the entire polynomial hierarchy
(Toda's theorem, 1989).
\end{remark}

\subsubsection{The Permanent--Determinant Gap}

\begin{theorem}[P\'olya, 1913]\label{thm:polya}
There is no $\pm 1$ matrix $D$ obtained from a $0/1$ matrix $A$ by
sign substitution such that $\det(D) = \mathrm{perm}(A)$ for all $A$,
when $A$ has size $n \ge 3$.
\end{theorem}

\begin{theorem}[Kasteleyn, 1961]\label{thm:kasteleyn}
For a planar graph $G$, the number of perfect matchings can be
computed in polynomial time via the FKT \emph{Pfaffian} algorithm:
choose a special $\pm 1$ orientation (\emph{Kasteleyn orientation})
of the edges so that the Pfaffian of the resulting skew-symmetric
matrix equals the matching count.
\end{theorem}

\begin{remark}
Kasteleyn's algorithm is the historical seed of \emph{holographic
algorithms} (Valiant, 2008): a meta-technique that maps certain
\#P-hard counting problems to determinants/Pfaffians via tensor
contractions on planar graphs.
\end{remark}

\subsubsection{Approximate Counting: JSV's FPRAS}

\begin{theorem}[Jerrum--Sinclair--Vigoda, 2004]\label{thm:jsv}
There is a fully polynomial randomized approximation scheme (FPRAS)
for the permanent of any non-negative matrix: an algorithm that, given
$A \in \mathbb R_{\ge 0}^{n \times n}$ and $\varepsilon, \delta > 0$,
computes $\widehat P$ with
$\Pr[|\widehat P - \mathrm{perm}(A)| > \varepsilon \, \mathrm{perm}(A)] \le \delta$
in time $\mathrm{poly}(n, 1/\varepsilon, \log(1/\delta))$.
\end{theorem}

\begin{proof}[Proof sketch]
Construct a Markov chain on the union of the set of perfect matchings
and \emph{near-perfect} matchings (missing exactly one edge from each
side). Bound the spectral gap using a sophisticated canonical-paths
analysis. Use the chain to sample matchings nearly uniformly, then
estimate the permanent by sampling and reweighting along a chain of
auxiliary matrices.
\end{proof}

\subsubsection{Graph-Theoretic Synthesis}

The permanent is the bridge between three areas:
\begin{itemize}
  \item \textbf{Combinatorics}: counts matchings, $0/1$ tilings,
        Latin squares, and lattice tilings.
  \item \textbf{Linear algebra}: a multilinear form invariant under
        \emph{row} permutations but not column permutations.
  \item \textbf{Complexity}: the canonical \#P-complete problem,
        whose tractability boundary (planar / dense / non-negative)
        traces a frontier of computational complexity.
\end{itemize}
Modern research includes deep neural-network embeddings of permanent
estimation (Sumi--Vempala, 2022) and the Barvinok polynomial
interpolation method, which gives a quasipolynomial deterministic
algorithm for permanents of \emph{strictly positive} matrices.

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute $\mathrm{perm}(A)$ and $\det(A)$ for
$A = \begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$ and verify that
$\mathrm{perm}(A) = 1$ counts the unique perfect matching of $P_3$
(viewed as a bipartite graph $\{a_1, a_2\} \cup \{b_1, b_2\}$).
\end{exercise}

\begin{exercise}
Use Theorem~\ref{thm:permmatch} to count perfect matchings of $K_{3,3}$:
its biadjacency matrix is the all-ones $3 \times 3$ matrix $J_3$, so
$\mathrm{perm}(J_3) = 3! = 6$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Apply Ryser's formula (Theorem~\ref{thm:ryser}) to compute
$\mathrm{perm}(J_4)$ (the all-ones $4 \times 4$ matrix), verifying
the count $4! = 24$ for perfect matchings of $K_{4,4}$.
\end{exercise}

\begin{exercise}
For a planar bipartite graph $G$, sketch the construction of a
Kasteleyn orientation (Theorem~\ref{thm:kasteleyn}). Apply it to the
$2 \times 2$ grid graph and verify that the Pfaffian count matches
the direct enumeration of perfect matchings.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Show that computing $\mathrm{perm}(A) \mod 2$ for a $0/1$ matrix is
in $\mathsf P$ (it equals $\det(A) \mod 2$), but computing
$\mathrm{perm}(A) \mod 3$ is \#P-hard (Valiant). This sharp dichotomy
illustrates the depth of Theorem~\ref{thm:valiant}.
\end{exercise}

\begin{exercise}
Use the Bregman--Minc inequality
$\mathrm{perm}(A) \le \prod_i (r_i!)^{1/r_i}$ for a $0/1$ matrix
$A$ with row sums $r_i$, to bound the number of perfect matchings of
a $d$-regular bipartite graph by $(d!)^{n/d}$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the bipartite graph corresponding to the biadjacency matrix
$\begin{pmatrix} 1 & 1 & 0 \\ 0 & 1 & 1 \\ 1 & 0 & 1 \end{pmatrix}$
and enumerate its perfect matchings to verify the permanent is $2$.
\end{exercise}

\begin{exercise}
For the $3 \times 3$ grid bipartite graph (vertices = cells,
edges = horizontal/vertical adjacency), draw a Kasteleyn orientation
and verify the Pfaffian count $\#\text{perfect matchings} = 0$
(impossible because of the odd-vertex parity).
\end{exercise}

%====================================================================
\subsection{Weisfeiler--Leman, Color Refinement, and Graph Isomorphism}
%====================================================================

\paragraph{Graph-theoretic motivation.}
Two graphs $G$ and $H$ are \emph{isomorphic} if there is a bijection
of vertices preserving adjacency. Deciding graph isomorphism (GI) is
the classical problem of testing structural equality of graphs;
since 2015 it has been known to be solvable in
$\mathrm{quasipoly}(n) = n^{O((\log n)^c)}$ time (Babai), placing
it in a unique complexity-theoretic position between $\mathsf{P}$ and
\textsf{NP-hard}. The algebraic core of every modern GI algorithm
is the \emph{Weisfeiler--Leman} (WL) hierarchy: an iterative
colour-refinement procedure that captures graph invariants of
increasing strength.

\subsubsection{Color Refinement (1-WL)}

\begin{definition}[Color refinement / 1-WL]\label{def:cr}
Initialize each vertex of $G$ with the colour $c^0(v) = $ degree of
$v$. At iteration $t \ge 1$, set
\[
  c^t(v) \;=\; \mathrm{hash}\big(c^{t-1}(v),\, \{\!\{c^{t-1}(u) : u \in N(v)\}\!\}\big),
\]
where $\{\!\{ \cdot \}\!\}$ is a multiset and $\mathrm{hash}$ is any
injective encoding of (vertex-colour, neighbour-multiset) pairs.
Iterate until the partition of $V(G)$ by colours stabilises (in at
most $|V(G)|$ rounds).
\end{definition}

\begin{theorem}[1-WL test]\label{thm:1wl}
Two graphs $G, H$ are \textbf{1-WL distinguishable} if the colour
multisets of $G$ and $H$ differ at some iteration. If $G$ and $H$ are
1-WL indistinguishable, they have the same degree sequence and many
finer invariants---but they may still be non-isomorphic.
\end{theorem}

\begin{example}[1-WL fails to distinguish CFI graphs]
The \emph{Cai--F\"urer--Immerman} (1992) construction produces, for
every $k$, a pair of non-isomorphic $k$-regular graphs that are
indistinguishable by $(k-1)$-WL. In particular, for any fixed $k$,
$k$-WL is strictly weaker than graph isomorphism testing.
\end{example}

\subsubsection{Higher-Dimensional Weisfeiler--Leman}

\begin{definition}[$k$-dimensional WL]\label{def:kwl}
For $k \ge 2$, $k$-WL operates on \emph{$k$-tuples} of vertices.
Initialize $c^0(\vec v) = $ isomorphism type of the induced subgraph
on $\{v_1, \ldots, v_k\}$ (with labelled vertices). At iteration $t$,
\[
  c^t(\vec v) \;=\; \mathrm{hash}\big(c^{t-1}(\vec v),\, \big(\{\!\{c^{t-1}(\vec v[i \to w]) : w \in V\}\!\}\big)_{i = 1}^{k}\big),
\]
where $\vec v[i \to w]$ replaces the $i$-th coordinate of $\vec v$
with $w$. Iterate until stabilisation.
\end{definition}

\begin{theorem}[Cai--F\"urer--Immerman, 1992]\label{thm:cfi}
For every $k$ there exists a graph pair $(G_k, H_k)$ such that
$G_k \not\cong H_k$ but $k$-WL fails to distinguish them. Thus the
$k$-WL hierarchy is a \emph{strict} infinite hierarchy.
\end{theorem}

\subsubsection{Babai's Quasi-Polynomial Algorithm}

\begin{theorem}[Babai, 2015--2016]\label{thm:babai}
Graph isomorphism is solvable in time $n^{O((\log n)^c)}$ for some
absolute constant $c$.
\end{theorem}

\begin{remark}
Babai's algorithm uses a deep blend of group theory and the
Weisfeiler--Leman hierarchy. The key technical tool is a
``\emph{split-or-Johnson}'' procedure that recursively reduces
graph isomorphism either to a smaller instance via canonical
splittings or to an action on a Johnson graph (which reveals
hidden bipartite structure).
\end{remark}

\subsubsection{WL and Graph Neural Networks}

\begin{theorem}[Xu--Hu--Leskovec--Jegelka; Morris et al., 2019]\label{thm:gnnwl}
The expressive power of standard message-passing graph neural networks
(GNNs) for distinguishing graphs is at most that of 1-WL: if 1-WL
cannot distinguish $G$ from $H$, then no GNN with the same number of
layers can output different graph-level representations on $G$ and
$H$.
\end{theorem}

\begin{remark}
Theorem~\ref{thm:gnnwl} sparked an explosion of \emph{higher-order
GNN} architectures (e.g.\ $k$-GNN, PPGN, LRGB) explicitly designed
to overcome the 1-WL bottleneck. The resulting GNN--WL correspondence
is now standard background in graph machine learning.
\end{remark}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Apply 1-WL to the cycle $C_4$ and to the disjoint union of two $C_3$'s.
Confirm that 1-WL produces different colour multisets, distinguishing
the graphs.
\end{exercise}

\begin{exercise}
Show that 1-WL cannot distinguish the cycle $C_6$ from the disjoint
union of two $C_3$'s. (Both are $2$-regular with the same number of
vertices.)
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Construct a pair of $3$-regular non-isomorphic graphs on $n = 8$
vertices that are 1-WL indistinguishable. (Hint: use a small
CFI-style construction.)
\end{exercise}

\begin{exercise}
Prove that $k$-WL is at least as strong as $(k - 1)$-WL: any pair
distinguished by $(k-1)$-WL is also distinguished by $k$-WL.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
State the precise CFI gadget construction (Theorem~\ref{thm:cfi}) and
verify that $k$-WL on the resulting graphs does not distinguish them.
\end{exercise}

\begin{exercise}
Prove that 1-WL is equivalent to vertex colour-refinement
\emph{in the sense of equitable partitions}: the stable colouring of
1-WL on $G$ corresponds to the coarsest equitable partition of $V(G)$.
Connect this to the spectrum of the quotient matrix from \S 12.2.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the trace of 1-WL on the Petersen graph: at each iteration, label
each vertex with its current colour. Observe that the algorithm
stabilises in $\le 2$ iterations and confirms vertex-transitivity.
\end{exercise}

\begin{exercise}
For a small CFI gadget on $4$ vertices, draw both graphs and label
the vertices with their 1-WL colours after stabilisation. Verify
that 1-WL fails to distinguish them.
\end{exercise}

%====================================================================
\subsection{Cluster Algebras from Quivers}
%====================================================================

\paragraph{Graph-theoretic motivation.}
A \emph{quiver} is just a directed graph (with multiple edges allowed
but no self-loops or oriented $2$-cycles). Out of any such graph,
Fomin and Zelevinsky (2002) conjured an algebraic structure---a
\emph{cluster algebra}---generated by recursive applications of a
graph operation called \emph{mutation}. Cluster algebras have since
appeared in dozens of mathematical fields: total positivity in Lie
groups, Teichm\"uller theory of surfaces, Calabi--Yau categories,
scattering amplitudes in physics, and the discrete dynamics of $Y$-
and $T$-systems. This subsection treats only the combinatorial heart:
mutation as a graph operation, and the Fomin--Zelevinsky finite-type
classification.

\subsubsection{Quivers, Skew-Symmetric Matrices, and Mutation}

\begin{definition}[Quiver]\label{def:quiver}
A \textbf{quiver} $Q$ is a directed multigraph with vertex set
$Q_0 = [n]$ and (multi)set of arrows $Q_1$. We assume $Q$ has no
loops and no oriented $2$-cycles. Encode $Q$ by its
\textbf{exchange matrix} $B = B(Q) \in \mathbb Z^{n \times n}$ with
$B_{ij} = $ (number of arrows $i \to j$) $-$ (number of arrows $j \to i$).
$B$ is skew-symmetric.
\end{definition}

\begin{definition}[Mutation at vertex $k$]\label{def:mutation}
The \textbf{mutation} $\mu_k$ of an exchange matrix $B$ produces a
new matrix $B'$ with entries
\[
  B'_{ij} \;=\;
  \begin{cases}
    -B_{ij} & \text{if } i = k \text{ or } j = k, \\
    B_{ij} + \frac{1}{2}\big(|B_{ik}| B_{kj} + B_{ik} |B_{kj}|\big) & \text{otherwise}.
  \end{cases}
\]
Equivalently, in terms of the quiver: reverse all arrows incident to
vertex $k$, and for every directed path $i \to k \to j$ add a new
arrow $i \to j$ (cancelling against any existing $j \to i$ arrows).
$\mu_k^2 = \mathrm{id}$.
\end{definition}

\begin{example}[Mutation on a triangle quiver]
Consider the quiver on $\{1, 2, 3\}$ with arrows $1 \to 2, 2 \to 3,
3 \to 1$ (an oriented $C_3$). After mutating at vertex $1$:
arrows $1 \to 2$ and $3 \to 1$ are reversed; the path $3 \to 1 \to 2$
contributes a new arrow $3 \to 2$, which cancels the existing
$2 \to 3$. Result: the quiver with arrows $2 \to 1, 1 \to 3$, and
nothing between $2$ and $3$ (i.e., $A_3$ orientation).
\end{example}

\subsubsection{Cluster Algebras}

\begin{definition}[Cluster algebra of a quiver]\label{def:cluster}
Fix variables $x_1, \ldots, x_n$ (an \emph{initial cluster}) and a
quiver $Q$ on $[n]$. The cluster algebra $\mathcal A(Q)$ is the
$\mathbb Z[x_1^{\pm 1}, \ldots, x_n^{\pm 1}]$-subalgebra of the field
$\mathbb Q(x_1, \ldots, x_n)$ generated by all \emph{cluster variables}
obtained from $(x_1, \ldots, x_n; Q)$ by iterated mutation, where
mutation at $k$ replaces $x_k$ by
\[
  x'_k \;=\; \frac{1}{x_k}\Big(\prod_{i : i \to k} x_i^{[B_{ik}]_+} + \prod_{j : k \to j} x_j^{[-B_{jk}]_+}\Big).
\]
\end{definition}

\begin{theorem}[Laurent phenomenon; Fomin--Zelevinsky, 2002]\label{thm:laurent}
Every cluster variable obtained by iterated mutation is a Laurent
polynomial in the initial variables $x_1, \ldots, x_n$ with integer
coefficients.
\end{theorem}

\begin{theorem}[Positivity; Lee--Schiffler, Gross--Hacking--Keel--Kontsevich, 2018]
The Laurent polynomials of Theorem~\ref{thm:laurent} have
\textbf{non-negative} integer coefficients.
\end{theorem}

\subsubsection{Finite-Type Classification}

\begin{theorem}[Fomin--Zelevinsky, 2003]\label{thm:fzfinite}
A cluster algebra $\mathcal A(Q)$ has only finitely many cluster
variables if and only if the quiver $Q$ is mutation-equivalent to an
orientation of a simply-laced Dynkin diagram of type
$A_n, D_n, E_6, E_7, E_8$.
\end{theorem}

\begin{remark}
The total number of cluster variables in a finite-type cluster algebra
depends only on the Dynkin type:
$|A_n| \to \binom{n + 2}{2} - 1, |D_n| \to n^2, |E_6| \to 42, |E_7| \to 70, |E_8| \to 128$.
This is the same count as the number of \emph{positive roots} plus
\emph{rank}, anticipating the deep connection between cluster algebras
and Lie theory.
\end{remark}

\subsubsection{Triangulations of Surfaces}

\begin{theorem}[Fomin--Shapiro--Thurston, 2008]\label{thm:fst}
The cluster algebra of a triangulated marked surface is encoded by
the \emph{adjacency quiver} of the triangulation: vertices = arcs;
arrows = consecutive pairs of arcs in a triangle. Flips of the
triangulation correspond exactly to mutations of the quiver.
\end{theorem}

\begin{example}[Hexagon triangulations]
For the hexagon (the disk with $6$ marked boundary points), the
cluster algebra is of type $A_3$. Its $14$ triangulations correspond
to the $14$ clusters; flips between them realise mutation.
\end{example}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Verify $\mu_k^2 = \mathrm{id}$ on the exchange matrix
$B = \begin{pmatrix} 0 & 1 \\ -1 & 0 \end{pmatrix}$ at vertex $k = 1$.
\end{exercise}

\begin{exercise}
Compute all clusters of the rank-$2$ cluster algebra of type $A_2$
(quiver: $1 \to 2$). The five clusters are $\{x_1, x_2\}, \{x_1, (1+x_1)/x_2\}, \ldots$
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Verify the Laurent phenomenon for the cluster algebra of type $A_2$:
all $5$ cluster variables are Laurent polynomials in $x_1, x_2$ with
non-negative integer coefficients.
\end{exercise}

\begin{exercise}
Prove that mutation $\mu_k$ on a quiver with no oriented $2$-cycles
again produces a quiver with no oriented $2$-cycles, given that the
formula resolves cancellations correctly.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove that any two adjacent triangulations of a polygon (differing
by one diagonal flip) correspond to clusters related by a single
mutation, completing the bijection of Theorem~\ref{thm:fst} for the
disk.
\end{exercise}

\begin{exercise}
Show that the mutation orbit of an orientation of $A_n$ is finite
(indeed, exactly $\binom{2(n+1)}{n+1}/(n+2)$ Catalan-many distinct
quivers up to relabeling), confirming Theorem~\ref{thm:fzfinite} for
type $A$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the quiver mutation graph (cluster exchange graph) of type $A_2$:
$5$ vertices, each connected to its two neighbours by an edge labeled
with the mutation index. Verify it is the pentagon.
\end{exercise}

\begin{exercise}
Sketch a triangulation of the hexagon with $3$ diagonals and the
corresponding adjacency quiver. Apply a flip to a chosen diagonal and
verify the resulting quiver is the mutation of the original at the
corresponding vertex.
\end{exercise}

%====================================================================
\subsection{Chip-Firing, Sandpile Groups, and Riemann--Roch for Graphs}
%====================================================================

\paragraph{Graph-theoretic motivation.}
On any finite graph, a deceptively simple game called \emph{chip
firing} (or the \emph{abelian sandpile model}) gives rise to a
discrete dynamical system whose orbit space is a finite abelian
group---the \emph{sandpile group} (also called the \emph{Picard
group}, \emph{critical group}, or \emph{Jacobian}). Baker and
Norine (2007) discovered that this group satisfies a graph-theoretic
analogue of the \emph{Riemann--Roch theorem} for divisors on a
Riemann surface. Chip firing thus realises a small but rigorous
version of intersection theory in pure combinatorics, and bridges
graph theory with algebraic geometry, tropical geometry, and the
theory of metric graphs.

\subsubsection{Divisors and Chip Firing}

\begin{definition}[Divisor]\label{def:divisor}
Let $G$ be a connected graph with vertex set $V$. A \textbf{divisor}
on $G$ is an element of the free abelian group
$\mathrm{Div}(G) = \mathbb Z^{V}$. Write a divisor as
$D = \sum_v D(v) \cdot v$. The \textbf{degree} of $D$ is
$\deg(D) = \sum_v D(v)$.
\end{definition}

\begin{definition}[Chip-firing move]\label{def:firing}
A \textbf{firing} at vertex $v$ replaces $D(v)$ by $D(v) - \deg_G(v)$
and increases $D(u)$ by the number of edges between $u$ and $v$ for
each $u \ne v$. In matrix language: $D \to D - L_v$, where $L_v$ is
the row of the graph Laplacian $L = D - A$ at vertex $v$. Fired
vertices preserve the total chip count $\deg(D)$.
\end{definition}

\begin{definition}[Linear equivalence]
Two divisors $D_1, D_2$ are \textbf{linearly equivalent}
($D_1 \sim D_2$) if $D_1 - D_2$ lies in the row span of $L$. The
quotient $\mathrm{Pic}(G) = \mathrm{Div}(G) / \sim$ is the
\textbf{Picard group} of $G$.
\end{definition}

\subsubsection{The Sandpile Group}

\begin{theorem}[Sandpile group structure]\label{thm:sandpile}
For a connected graph $G$ on $n$ vertices,
$\mathrm{Pic}(G) \cong \mathbb Z \oplus K(G)$, where $K(G)$, the
\textbf{sandpile group}, is a finite abelian group with
$|K(G)| = \tau(G)$, the number of spanning trees.
\end{theorem}

\begin{example}[Sandpile group of small graphs]
\mbox{}
\begin{itemize}
  \item $K(K_n) \cong (\mathbb Z / n\mathbb Z)^{n - 2}$ (so
        $|K(K_n)| = n^{n - 2}$, recovering Cayley's formula).
  \item $K(C_n) \cong \mathbb Z / n \mathbb Z$.
  \item $K(K_{m, n}) \cong (\mathbb Z / m \mathbb Z)^{n - 1} \oplus (\mathbb Z / n \mathbb Z)^{m - 1}$.
\end{itemize}
\end{example}

\subsubsection{The Baker--Norine Riemann--Roch Theorem}

\begin{definition}[Effective divisor; rank]
A divisor $D$ is \textbf{effective} ($D \ge 0$) if $D(v) \ge 0$ for
all $v$. The \textbf{rank} of $D$ is $r(D) = -1$ if no effective
divisor is linearly equivalent to $D$, and otherwise
\[
  r(D) \;=\; \max\{\, k \ge 0 :\ D - E \sim E' \ge 0 \text{ for every effective } E \text{ of degree } k\,\}.
\]
\end{definition}

\begin{definition}[Canonical divisor]
The \textbf{canonical divisor} of $G$ is $K_G = \sum_v (\deg_G(v) - 2) \cdot v$.
Its degree is $2g$ where $g = |E| - |V| + 1$ is the cycle rank
(\textbf{genus}) of $G$.
\end{definition}

\begin{theorem}[Baker--Norine, 2007]\label{thm:bakernorine}
For any connected graph $G$ and any divisor $D$,
\[
  r(D) - r(K_G - D) \;=\; \deg(D) - g + 1.
\]
\end{theorem}

\begin{remark}
Theorem~\ref{thm:bakernorine} is a literal Riemann--Roch theorem:
it has the same form as Riemann--Roch on a smooth projective curve,
with the cycle rank $g$ playing the role of geometric genus. The
proof is purely combinatorial (no algebraic geometry required), but
the deep parallels with algebraic curves were the spark for
\textbf{tropical geometry} of graphs (Mikhalkin, Baker--Jensen,
Caporaso--Viviani).
\end{remark}

\subsubsection{Graph-Theoretic Connections}

\begin{itemize}
  \item \textbf{Spanning trees}: $|K(G)| = \tau(G)$ recovers
        Kirchhoff (\S 11.4) by the Matrix-Tree theorem applied to
        the Laplacian.
  \item \textbf{Counting recurrent configurations}: the sandpile
        group elements are in bijection with \emph{recurrent}
        chip-firing configurations, themselves counted by the Tutte
        polynomial $T_G(1, y)$.
  \item \textbf{Random spanning trees}: Wilson's algorithm
        (\emph{loop-erased random walk}) generates a uniform spanning
        tree in expected time $O(\tau(G)/|E|)$.
  \item \textbf{$g$-theorem for graphs}: Riemann--Roch implies the
        \emph{Clifford bound} $r(D) \le \deg(D)/2$ for divisors of
        degree $\le 2g - 2$, the analogue of the Clifford theorem on
        algebraic curves.
\end{itemize}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute the sandpile group of $C_4$. (Answer: $\mathbb Z / 4 \mathbb Z$,
matching $\tau(C_4) = 4$.)
\end{exercise}

\begin{exercise}
For $K_4$, compute the canonical divisor $K_{K_4}$ and verify
$\deg(K_{K_4}) = 2g$ where $g = 6 - 4 + 1 = 3$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Verify Theorem~\ref{thm:sandpile} for $K_3$: compute $L$ explicitly,
identify the sandpile group as $\mathbb Z / 3 \mathbb Z$, and check
that $|K(K_3)| = 3 = \tau(K_3)$.
\end{exercise}

\begin{exercise}
Apply the Baker--Norine theorem to a divisor of degree $g$ on $K_4$
and verify $r(D) - r(K_G - D) = \deg(D) - g + 1 = 1$ holds for at
least one such $D$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove the matrix-tree count $|K(G)| = \tau(G)$ from
Theorem~\ref{thm:sandpile}: the Smith normal form of the reduced
Laplacian gives both the abelian-group structure and the determinant
$\det(L_{\text{reduced}}) = \tau(G)$.
\end{exercise}

\begin{exercise}
Sketch a proof of the Baker--Norine Riemann--Roch theorem
(\ref{thm:bakernorine}) via Dhar's burning algorithm and the
\emph{$q$-reduced} divisor canonical form.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw a chip-firing trajectory on $K_4$ starting from
$D_0 = 5 v_1 + 0 v_2 + 0 v_3 + 0 v_4$ until the configuration
stabilises. Identify the resulting divisor as a representative of an
element of $K(K_4)$.
\end{exercise}

\begin{exercise}
For the complete bipartite graph $K_{2, 3}$, draw all $6$ spanning
trees and verify $\tau = 6 = |K(K_{2,3})|$ by computing the sandpile
group explicitly via Smith normal form.
\end{exercise}

%====================================================================
\subsection{Random Walks via Representation Theory}
%====================================================================

\paragraph{Graph-theoretic motivation.}
Theorem~\ref{thm:cayleyabelian} computed eigenvalues of abelian
Cayley graphs from characters; the same idea, generalised to
non-abelian groups, gives the sharpest known mixing-time bounds for
random walks on highly symmetric graphs. The
\emph{Diaconis--Shahshahani upper bound lemma} packages this into a
single inequality controlling total-variation distance to stationarity
in terms of representation-theoretic data. The technique is the
modern engine for bounding mixing times of card-shuffling, lattice
walks, and the Markov chains used in the FPRAS algorithms of
\S 12.10.

\subsubsection{The Upper Bound Lemma}

\begin{definition}[Convolution Markov chain]\label{def:convchain}
Let $\Gamma$ be a finite group and $\mu : \Gamma \to \mathbb R$ a
probability measure. The \textbf{convolution random walk} on $\Gamma$
driven by $\mu$ has transition kernel $P(g, h) = \mu(g^{-1} h)$. After
$k$ steps, the law is $\mu^{*k}$ (the $k$-fold convolution power).
\end{definition}

\begin{theorem}[Diaconis--Shahshahani upper bound lemma, 1981]\label{thm:dsubl}
Let $\Gamma$ be a finite group with normalised Haar measure $U$ and
$\mu$ a symmetric probability measure on $\Gamma$ supported on a
generating set. For every $k \ge 0$,
\[
  4 \big\| \mu^{*k} - U \big\|_{\mathrm{TV}}^2
  \;\le\;
  \sum_{\rho \ne \mathbf{1}} \dim(\rho)\, \big\|\widehat\mu(\rho)^k\big\|_F^2,
\]
where the sum is over non-trivial irreducible unitary
representations $\rho$ of $\Gamma$, $\widehat\mu(\rho) = \sum_g \mu(g) \rho(g)$
is the Fourier transform of $\mu$ at $\rho$, and $\| \cdot \|_F$ is
the Frobenius norm.
\end{theorem}

\subsubsection{Worked Examples}

\begin{example}[Random transposition walk on $S_n$; Diaconis--Shahshahani, 1981]
The walk on $S_n$ driven by the uniform measure on transpositions
satisfies
\[
  \big\| \mu^{*k} - U \big\|_{\mathrm{TV}}
  \;=\;
  \begin{cases}
    1 - o(1) & \text{if } k = \tfrac{1}{2} n \log n - c n, \\
    o(1) & \text{if } k = \tfrac{1}{2} n \log n + c n,
  \end{cases}
\]
exhibiting a sharp \textbf{cutoff} at time $\tfrac{1}{2} n \log n$.
The bound uses Frobenius's formula for character values on
transpositions in terms of Young-diagram contents.
\end{example}

\begin{example}[Riffle shuffle; Bayer--Diaconis, 1992]
The Gilbert--Shannon--Reeds riffle shuffle on $52$ cards exhibits
cutoff at time $\tfrac{3}{2} \log_2 52 \approx 8.5$ shuffles, justifying
the famous ``seven shuffles'' rule of thumb.
\end{example}

\begin{example}[Hypercube random walk]
The lazy random walk on $Q_n$ (driven by the measure putting equal
mass on $e_1, \ldots, e_n$ and a self-loop) has eigenvalues
$\lambda_k = (1 + (1 - 2k/n))/2$. The mixing time is
$t_{\mathrm{mix}} = \tfrac{1}{2} n \log n + c n$, with cutoff at this
time.
\end{example}

\subsubsection{Cutoff Phenomenon}

\begin{definition}[Cutoff]\label{def:cutoff}
A family of Markov chains $\{P_n\}$ exhibits \textbf{cutoff} at
times $t_n$ with window $w_n$ if for every $\varepsilon > 0$,
\[
  \lim_{n \to \infty} d_{\mathrm{TV}}(P_n^{t_n - C w_n}, U_n) = 1
  \quad\text{and}\quad
  \lim_{n \to \infty} d_{\mathrm{TV}}(P_n^{t_n + C w_n}, U_n) = 0
\]
for sufficiently large $C = C(\varepsilon)$.
\end{definition}

\begin{theorem}[Cutoff for transitive expanders; Lubetzky--Sly, 2011]
Every sequence of vertex-transitive graphs satisfying a uniform
Cheeger expansion bound exhibits cutoff for the lazy simple random
walk, with cutoff time $\log n / \log(1/\lambda)$ where $\lambda$ is
the second-largest absolute eigenvalue.
\end{theorem}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
For the lazy random walk on $C_n$, verify the eigenvalues $\lambda_k = (1 + \cos(2\pi k / n))/2$
and bound the mixing time by $O(n^2)$.
\end{exercise}

\begin{exercise}
Apply Theorem~\ref{thm:dsubl} to the lazy random walk on $\mathbb Z_2^n$
(driven by uniform measure on standard basis vectors). Use the
character formula to recover the cutoff time $\tfrac{1}{2} n \log n$.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
For the random transposition walk on $S_n$, compute the first-order
upper bound on TV-distance using only the trivial representation, the
sign representation, and the standard $(n-1)$-dimensional
representation.
\end{exercise}

\begin{exercise}
Prove the cutoff phenomenon for the lazy random walk on the hypercube
$Q_n$ at time $\tfrac{1}{2} n \log n$, using the Diaconis--Shahshahani
lemma and the explicit eigenvalues.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
State and prove the Diaconis--Shahshahani upper bound lemma
(Theorem~\ref{thm:dsubl}) using the Plancherel formula on $\Gamma$.
\end{exercise}

\begin{exercise}
Construct an explicit family of $3$-regular Cayley graphs on
$\mathrm{PSL}(2, p)$ (the Lubotzky--Phillips--Sarnak Ramanujan graphs)
and prove they have spectral gap $\Omega(1)$ uniformly in $p$.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Sketch the TV-distance to stationarity as a function of step number
$k$ for the lazy random walk on $Q_5$. Highlight the cutoff window
around $k = \tfrac{1}{2} \cdot 5 \cdot \log 5 \approx 4$ steps.
\end{exercise}

\begin{exercise}
Draw the Cayley graph of $S_3$ generated by $\{(12), (23)\}$ (the
hexagon), and trace the first three steps of the random walk from the
identity. Use the explicit non-abelian character formula to compute
$\mu^{*1}, \mu^{*2}, \mu^{*3}$.
\end{exercise}

%====================================================================
\subsection{Electrical Networks and Effective Resistance}
%====================================================================

\paragraph{Graph-theoretic motivation.}
Treat each edge of a graph as a unit resistor; the resulting
electrical network is dual to the simple random walk. Many
graph-theoretic invariants---spanning tree counts, hitting times,
commute times, the Laplacian pseudoinverse---admit clean
expressions in terms of \emph{effective resistance}. This is one of
the most useful and least technical bridges between linear algebra
and graph theory; the standard references are Doyle--Snell's classic
\emph{Random Walks and Electric Networks} (1984) and the modern
treatment in Lyons--Peres, \emph{Probability on Trees and Networks}
(2016).

\subsubsection{Effective Resistance}

\begin{definition}[Effective resistance]\label{def:efftr}
Let $G$ be a connected graph with each edge assigned conductance $1$.
The \textbf{effective resistance} between vertices $u, v \in V(G)$,
denoted $R_{\mathrm{eff}}(u, v)$, is the voltage drop induced by
injecting $1$ unit of current at $u$ and extracting it at $v$, with
the resulting voltage function $\phi$ chosen to minimise the
quadratic form $\sum_{e = ij \in E} (\phi_i - \phi_j)^2$.
Equivalently, $R_{\mathrm{eff}}(u, v) = (e_u - e_v)^\top L^+ (e_u - e_v)$
where $L^+$ is the Moore--Penrose pseudoinverse of the Laplacian.
\end{definition}

\begin{theorem}[Foster's network theorem, 1949]\label{thm:foster}
For a connected graph $G$ on $n$ vertices,
\[
  \sum_{e = ij \in E(G)} R_{\mathrm{eff}}(i, j) \;=\; n - 1.
\]
\end{theorem}

\begin{theorem}[Kirchhoff's two-laws and the Laplacian]
The voltage function $\phi$ for unit current from $u$ to $v$ is the
unique solution (up to additive constant) to $L \phi = e_u - e_v$.
The current flowing along edge $ij$ in the direction $i \to j$ is
$\phi_i - \phi_j$. Energy dissipation
$\sum_{ij} (\phi_i - \phi_j)^2 = R_{\mathrm{eff}}(u, v)$.
\end{theorem}

\subsubsection{Spanning Trees and Resistance}

\begin{theorem}[Kirchhoff's effective resistance / spanning tree formula]
\label{thm:effspan}
For an edge $e = ij$ in $G$,
\[
  R_{\mathrm{eff}}(i, j) \;=\; \frac{\tau(G/e)}{\tau(G)},
\]
where $G/e$ is the contraction of $e$ and $\tau$ counts spanning
trees. Equivalently, the probability that edge $e$ appears in a
uniformly random spanning tree is $R_{\mathrm{eff}}(i, j)$.
\end{theorem}

\subsubsection{Hitting and Commute Times}

\begin{definition}[Hitting time, commute time]
For the simple random walk on $G$, $H(u, v) = $ expected number of
steps for the walk started at $u$ to first reach $v$. The
\textbf{commute time} is $C(u, v) = H(u, v) + H(v, u)$.
\end{definition}

\begin{theorem}[Chandra--Raghavan--Ruzzo--Smolensky--Tiwari, 1996]\label{thm:crrst}
\[
  C(u, v) \;=\; 2 |E| \cdot R_{\mathrm{eff}}(u, v).
\]
\end{theorem}

\begin{remark}
Theorem~\ref{thm:crrst} translates between random-walk and electrical
quantities exactly, and is the basis of the \emph{spectral sparsifier}
algorithms of Spielman--Teng (2004): a graph $G$ can be approximated
in nearly-linear time by a sparse graph $H$ such that
$L_G \approx L_H$ in the L\"owner sense, and consequently
$R_{\mathrm{eff}}^G \approx R_{\mathrm{eff}}^H$ everywhere.
\end{remark}

\subsubsection{Schramm--Lov\'asz Circle Packings}

\begin{theorem}[Schramm--Lov\'asz, 1990s]
Every $3$-connected planar graph admits an essentially unique
\emph{circle packing}: each vertex corresponds to a circle in the
plane, with two circles externally tangent iff the vertices are
adjacent. The Cheeger constant of the graph is comparable to the
\emph{conformal modulus} of the corresponding packing, providing a
geometric route to spectral gaps.
\end{theorem}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute $R_{\mathrm{eff}}$ between adjacent vertices of $C_4$. Verify
Foster's theorem: $4 \cdot R_{\mathrm{eff}} = 3$.
\end{exercise}

\begin{exercise}
For the path $P_n$, derive $R_{\mathrm{eff}}(1, n) = n - 1$ and verify
Foster's theorem.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
For the Petersen graph, compute $R_{\mathrm{eff}}$ between adjacent
vertices and between non-adjacent vertices using its spectrum
$\{3, 1^{(5)}, -2^{(4)}\}$ and the Laplacian-pseudoinverse formula.
\end{exercise}

\begin{exercise}
Verify Theorem~\ref{thm:effspan} for $K_4$: compute $\tau(K_4) = 16$
and $\tau(K_4 / e) = ?$ for an edge $e$, and check the formula
$R_{\mathrm{eff}}(e) = \tau(K_4 / e) / \tau(K_4)$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove Foster's network theorem (\ref{thm:foster}) by computing
$\mathrm{tr}(L L^+) = n - 1$ via the trace identity.
\end{exercise}

\begin{exercise}
Prove the Chandra--Raghavan formula $C(u, v) = 2|E| R_{\mathrm{eff}}(u, v)$
by relating both quantities to harmonic functions on $G \setminus \{v\}$
and the Laplacian pseudoinverse.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the circle packing of the cube graph $Q_3$ on the sphere
(stereographically projected to the plane). Verify there are $8$
circles, with adjacencies matching the edges.
\end{exercise}

\begin{exercise}
Sketch a current flow of $1$ ampere from one vertex of $K_4$ to the
opposite vertex. Label each edge with the current $\phi_i - \phi_j$.
Verify Kirchhoff's current law at each internal vertex.
\end{exercise}

%====================================================================
\subsection{Crystal Graphs and the Plactic Monoid}
%====================================================================

\paragraph{Graph-theoretic motivation.}
A \emph{crystal} is an edge-labelled directed graph (introduced by
Kashiwara, 1990) whose connected components encode the irreducible
representations of a Lie algebra. For type $A$ ($\mathfrak{gl}_n$),
crystal graphs are explicitly the graphs on \emph{semistandard
Young tableaux}, with edges given by Kashiwara operators
$\widetilde e_i, \widetilde f_i$. The combinatorial structure of these
graphs encodes the Schur expansion of products of Schur functions
(the Littlewood--Richardson rule) and recovers the Robinson--Schensted
correspondence as a graph homomorphism. This subsection is the most
algebraic in the chapter; we develop only enough crystal theory to
define the plactic monoid and connect it back to graph theory.

\subsubsection{Crystals of Type $A_{n-1}$}

\begin{definition}[Crystal]\label{def:crystal}
A \textbf{type-$A_{n-1}$ crystal} is a finite set $B$ equipped with
maps $\widetilde e_i, \widetilde f_i : B \to B \cup \{0\}$ for
$i = 1, \ldots, n - 1$, and weight function $\mathrm{wt} : B \to \mathbb Z^n$,
satisfying:
\begin{enumerate}
  \item $\widetilde f_i(b) = b'$ iff $\widetilde e_i(b') = b$
        (the operators are partial inverses).
  \item $\mathrm{wt}(\widetilde f_i b) = \mathrm{wt}(b) - \alpha_i$
        where $\alpha_i = e_i - e_{i+1}$ is the simple root.
  \item Standard string-length axioms.
\end{enumerate}
The \textbf{character} of $B$ is $\mathrm{ch}(B) = \sum_{b \in B} x^{\mathrm{wt}(b)}$.
\end{definition}

\begin{theorem}[Crystal of a polynomial representation]\label{thm:crysrep}
For each partition $\lambda$ with at most $n$ parts, there is a
unique connected crystal $B(\lambda)$ on the semistandard Young
tableaux of shape $\lambda$ over the alphabet $[n]$, with character
\[
  \mathrm{ch}(B(\lambda)) \;=\; s_\lambda(x_1, \ldots, x_n),
\]
the Schur polynomial.
\end{theorem}

\subsubsection{The Plactic Monoid}

\begin{definition}[Knuth equivalence]
On the free monoid $[n]^*$ (words over $\{1, \ldots, n\}$), define the
\textbf{Knuth equivalence} $\sim_K$ as the congruence generated by
\begin{itemize}
  \item $acb \sim_K cab$ if $a \le b < c$,
  \item $bac \sim_K bca$ if $a < b \le c$.
\end{itemize}
\end{definition}

\begin{definition}[Plactic monoid]
The \textbf{plactic monoid} $\mathcal P_n$ is $[n]^* / \sim_K$, with
multiplication inherited from concatenation.
\end{definition}

\begin{theorem}[Lascoux--Sch\"utzenberger, 1981]\label{thm:plactic}
Knuth equivalence classes are in bijection with semistandard Young
tableaux: each word $w$ has a unique tableau $P(w)$ such that
$w \sim_K \mathrm{row}(P(w))$ (the row reading word). Hence
$\mathcal P_n \cong \bigsqcup_{\lambda} \mathrm{SSYT}(\lambda, n)$ as
sets, and the multiplication corresponds to the
\emph{Schensted insertion} algorithm.
\end{theorem}

\subsubsection{Robinson--Schensted as a Graph Homomorphism}

\begin{theorem}[RSK and crystal graphs]
The Robinson--Schensted correspondence $\pi \mapsto (P, Q)$ for
permutations $\pi \in S_n$ is realised by the morphism of crystals
$B(1)^{\otimes n} \to \bigsqcup_\lambda B(\lambda) \otimes B(\lambda)$
given by Knuth-equivalence-class extraction. The shape $\lambda$ is
the connected component containing $\pi$.
\end{theorem}

\subsubsection{Application: Littlewood--Richardson Rule}

\begin{theorem}[Littlewood--Richardson via crystals]
The structure constants $c_{\mu \nu}^\lambda$ in
$s_\mu \cdot s_\nu = \sum_\lambda c_{\mu\nu}^\lambda s_\lambda$
equal the number of \textbf{Yamanouchi tableaux} of shape $\lambda / \mu$
and content $\nu$, equivalently, the number of crystal isomorphisms
$B(\lambda) \hookrightarrow B(\mu) \otimes B(\nu)$.
\end{theorem}

\subsubsection{Connection to Graph Theory}

The crystal graph of $B(\lambda)$ is an edge-labelled directed graph
on $\dim s_\lambda$ vertices. As a \emph{graph}:
\begin{itemize}
  \item Connectivity = irreducibility of the underlying representation.
  \item The graph's automorphism group is closely tied to the Weyl
        group of $\mathrm{GL}_n$.
  \item Counting paths in crystal graphs of products $B(\lambda) \otimes B(\mu)$
        recovers the Littlewood--Richardson coefficients.
  \item Specialisation to the standard alphabet recovers the
        \emph{RSK correspondence} bijection between permutations and
        pairs of standard tableaux---a foundational result that
        connects the symmetric group $S_n$ to combinatorial graph
        invariants like the longest increasing subsequence.
\end{itemize}

\subsubsection{Exercises}

\paragraph{Easy Questions.}

\begin{exercise}
Compute the crystal graph $B(1, 1)$ over the alphabet $\{1, 2, 3\}$:
the SSYT of shape $(1, 1)$ are columns $\binom{i}{j}$ with $i < j$,
giving $\binom{3}{2} = 3$ vertices. Draw the crystal with
$\widetilde f_1, \widetilde f_2$ edges.
\end{exercise}

\begin{exercise}
Apply Schensted insertion to the word $w = 3 1 4 1 5$ and confirm the
resulting tableau is the unique element of its Knuth equivalence
class.
\end{exercise}

\paragraph{Medium Questions.}

\begin{exercise}
Verify Theorem~\ref{thm:crysrep} for $\lambda = (2, 1)$ over $\{1, 2\}$:
the crystal has $\binom{3}{2} = 3$ tableaux, and its character
$\mathrm{ch}(B(2, 1)) = x_1^2 x_2 + x_1 x_2^2$ (degree-$3$ Schur with
two parts).
\end{exercise}

\begin{exercise}
Implement the Knuth-equivalence relations on $\{1, 2, 3\}^4$ and count
the equivalence classes. Verify the count against the dimension sum
$\sum_\lambda (f^\lambda)^2$ over partitions of $4$.
\end{exercise}

\paragraph{Hard Questions.}

\begin{exercise}
Prove the Littlewood--Richardson rule using crystal techniques: the
connected components of $B(\mu) \otimes B(\nu)$ correspond to
Yamanouchi tableaux of shape $\lambda / \mu$ and content $\nu$.
\end{exercise}

\begin{exercise}
Show that the longest increasing subsequence in a permutation $\pi$
equals the length of the first row of $P(\pi)$ (Schensted's theorem),
using the crystal-graph interpretation of the RSK correspondence.
\end{exercise}

\paragraph{Diagram Questions.}

\begin{exercise}
Draw the crystal graph $B(1, 1, 1)$ over $\{1, 2, 3\}$: the unique
SSYT of shape $(1, 1, 1)$ is the column $1, 2, 3$, so the crystal has
a single vertex. Verify $\mathrm{ch}(B) = x_1 x_2 x_3$.
\end{exercise}

\begin{exercise}
Sketch the crystal $B(2)$ over $\{1, 2, 3\}$ (single-row SSYT of
length $2$). It has $\binom{4}{2} = 6$ vertices; draw the
$\widetilde f_1, \widetilde f_2$ edges and verify the character
$h_2(x_1, x_2, x_3) = \sum_{i \le j} x_i x_j$.
\end{exercise}

%====================================================================
\subsection{Further Reading and Pointers}
%====================================================================

\paragraph{Matchings polynomial and rootedness.} Godsil's
\emph{Algebraic Combinatorics} (1993), Chapters 1, 4, 6, develops the
path-tree machinery in full and extends to the broader theory of
\emph{tree-like} walks and characteristic polynomials of covers.

\paragraph{Equitable partitions and walk-regularity.} Godsil--Royle,
\emph{Algebraic Graph Theory} (2001), Chapters 9--10, are the
canonical reference. Brouwer--Haemers, \emph{Spectra of Graphs}
(2012), Chapter 3, contains modern interlacing and Hoffman/Haemers
bounds.

\paragraph{Distance-regular graphs and association schemes.}
Brouwer--Cohen--Neumaier, \emph{Distance-Regular Graphs} (1989), is
the comprehensive reference; Bannai--Ito, \emph{Algebraic
Combinatorics I} (1984), develops the Bose--Mesner algebra. Delsarte's
1973 thesis on the LP method remains foundational.

\paragraph{Cayley graphs and expanders.} Hoory--Linial--Wigderson,
``Expander Graphs and Their Applications'' (\emph{Bull.\ AMS}, 2006),
surveys both deterministic and probabilistic constructions. Lubotzky,
\emph{Discrete Groups, Expanding Graphs, and Invariant Measures}
(1994), connects to representation theory and number theory.

\paragraph{Tutte polynomial and matroids.} Bollob\'as,
\emph{Modern Graph Theory} (1998), Chapter 10, gives a graph-theoretic
treatment. Oxley, \emph{Matroid Theory} (2nd ed., 2011), is the
matroid-side reference. Welsh's
\emph{Complexity: Knots, Colourings and Counting} (1993) discusses
the Jones polynomial connection.

\paragraph{Chromatic symmetric function.} Stanley's original paper,
``A symmetric function generalization of the chromatic polynomial of
a graph'' (\emph{Adv.\ Math.}, 1995), is short and self-contained.
Recent partial progress on the tree conjecture is surveyed in
Crew--Spirkl (2023) and references therein.

\paragraph{Heaps of pieces and walk generating functions.} Viennot's
\emph{Heaps of Pieces, I: Basic Definitions and Combinatorial
Lemmas} (1986) introduced the theory and is the canonical reference.
Garsia--Egecioglu, \emph{Lectures in Algebraic Combinatorics} (2020),
Chapter 4, gives a modern self-contained exposition with applications
to the matchings polynomial and orthogonal polynomials. Flajolet's
``Combinatorial aspects of continued fractions'' (\emph{Discrete
Math.}, 1980) is the source of the continued-fraction theorem
(Theorem~\ref{thm:flajolet}). For the Lindstr\"om--Gessel--Viennot
lemma in this language, see Stanley, \emph{Enumerative Combinatorics,
Vol.\ I} (2nd ed., 2012), \S 2.7.

\paragraph{Ihara zeta function.} Stark--Terras, \emph{Zeta functions
of finite graphs and coverings} (parts I--III, \emph{Adv.\ Math.}\
1996/2000/2007) develop the theory thoroughly. Terras's
\emph{Zeta Functions of Graphs: A Stroll through the Garden} (2010)
is the canonical book-length reference. For the non-backtracking
matrix in random graph theory and community detection, see
Krzakala--Moore--Mossel--Neeman--Sly--Zdeborov\'a--Zhang,
``Spectral redemption in clustering sparse networks'' (\emph{PNAS},
2013). Bass's three-term identity is in his 1992 paper
``The Ihara--Selberg zeta function of a tree lattice''.

\paragraph{Combinatorial Hopf algebras.} Schmitt's foundational paper
``Incidence Hopf algebras'' (\emph{J.\ Pure Appl.\ Algebra}, 1994)
introduces the graph and matroid Hopf algebras. Aguiar--Bergeron--Sottile,
``Combinatorial Hopf algebras and generalized Dehn--Sommerville
relations'' (\emph{Compos.\ Math.}, 2006), is the universal-property
theorem; Aguiar--Mahajan, \emph{Monoidal Functors, Species, and Hopf
Algebras} (2010), is a comprehensive modern treatment. For the
Stanley--Stembridge conjecture and Hikita's recent proof, see
Hikita's 2024 arXiv preprint and the survey by Shareshian--Wachs.
Connes--Kreimer, ``Hopf algebras, renormalization and noncommutative
geometry'' (\emph{Comm.\ Math.\ Phys.}, 1998), connects Hopf algebras
to QFT.

\paragraph{Hodge theory of matroids.} Adiprasito--Huh--Katz,
``Hodge theory for combinatorial geometries'' (\emph{Annals of Math.},
2018), is the main paper; the Bourbaki seminar of Bonifant--Huh
(2019) and the survey of Huh, ``Combinatorial applications of the
Hodge--Riemann relations'' (ICM, 2018), are excellent introductions.
For spectral independence and modern matroid sampling, see
Anari--Liu--Oveis Gharan--Vinzant, ``Log-concave polynomials II:
high-dimensional walks and an FPRAS for counting bases of a matroid''
(\emph{STOC}, 2019). The Feichtner--Yuzvinsky paper ``Chow rings of
toric varieties defined by atomic lattices'' (\emph{Invent.\ Math.},
2004) introduces the Chow ring construction.

\paragraph{Permanent and \#P.} Valiant's original paper
``The complexity of computing the permanent'' (\emph{Theoret.\ Comput.\ Sci.},
1979) defines \#P and proves Theorem~\ref{thm:valiant}. Minc's
\emph{Permanents} (1978) is the classical compendium of
permanent identities including Ryser's formula and the Bregman--Minc
inequality. For Kasteleyn's planar Pfaffian algorithm, see
Kasteleyn, ``Graph theory and crystal physics'' (in F.\ Harary, ed.,
\emph{Graph Theory and Theoretical Physics}, 1967). Jerrum--Sinclair--Vigoda,
``A polynomial-time approximation algorithm for the permanent of a
matrix with non-negative entries'' (\emph{J.\ ACM}, 2004), is the
FPRAS paper. Barvinok's \emph{Combinatorics and Complexity of
Partition Functions} (2017) covers polynomial interpolation
techniques.

\paragraph{Weisfeiler--Leman and graph isomorphism.} Cai--F\"urer--Immerman,
``An optimal lower bound on the number of variables for graph
identification'' (\emph{Combinatorica}, 1992), gives the foundational
hierarchy lower bound. Babai's quasipolynomial paper, ``Graph
isomorphism in quasipolynomial time'' (arXiv:1512.03547, 2015--16),
remains the state of the art. For the GNN--WL connection, see
Xu--Hu--Leskovec--Jegelka, ``How powerful are graph neural networks?''
(\emph{ICLR}, 2019), and Morris et al., ``Weisfeiler and Leman go
neural: higher-order graph neural networks'' (\emph{AAAI}, 2019).
Grohe's \emph{Descriptive Complexity, Canonisation, and Definable
Graph Structure Theory} (2017) is the canonical reference.

\paragraph{Cluster algebras.} Fomin--Zelevinsky,
``Cluster algebras I: Foundations'' (\emph{J.\ Amer.\ Math.\ Soc.},
2002), launched the field; the four-paper series concludes with the
finite-type classification. Marsh's \emph{Lecture Notes on Cluster
Algebras} (2014) is a friendly introduction. Fomin--Williams--Zelevinsky's
ongoing book \emph{Introduction to Cluster Algebras} (chapters posted
on arXiv since 2016) is the developing canonical reference.
Triangulations / surfaces: Fomin--Shapiro--Thurston,
``Cluster algebras and triangulated surfaces I: Cluster complexes''
(\emph{Acta Math.}, 2008). For positivity, Lee--Schiffler,
``Positivity for cluster algebras'' (\emph{Annals of Math.}, 2015),
and Gross--Hacking--Keel--Kontsevich,
``Canonical bases for cluster algebras'' (\emph{J.\ Amer.\ Math.\ Soc.},
2018).

\paragraph{Chip-firing and Riemann--Roch for graphs.} Baker--Norine,
``Riemann--Roch and Abel--Jacobi theory on a finite graph''
(\emph{Adv.\ Math.}, 2007), is the foundational paper. Klivans's
\emph{The Mathematics of Chip-Firing} (2018) is the book-length
treatment. Corry--Perkinson, \emph{Divisors and Sandpiles: An
Introduction to Chip-Firing} (2018), gives a graduate-level
exposition with many exercises. For tropical extensions,
Baker--Jensen, ``Degeneration of linear series from the tropical
point of view'' (\emph{Algebr.\ Geom.}, 2016).

\paragraph{Random walks via representation theory.} Diaconis,
\emph{Group Representations in Probability and Statistics}
(IMS Lecture Notes, 1988) is the canonical reference. The
random-transposition paper is Diaconis--Shahshahani, ``Generating a
random permutation with random transpositions''
(\emph{Z.\ Wahrsch.\ Verw.\ Gebiete}, 1981). Bayer--Diaconis,
``Trailing the dovetail shuffle to its lair''
(\emph{Ann.\ Appl.\ Probab.}, 1992), proves the riffle-shuffle cutoff.
Lubetzky--Sly, ``Cutoff phenomena for random walks on random regular
graphs'' (\emph{Duke Math.\ J.}, 2010), proves cutoff for random
regular graphs. Saloff-Coste, \emph{Lectures on finite Markov chains}
(\emph{\'Ecole d'\'Et\'e de Probabilit\'es de Saint-Flour XXVI},
1996) is also highly recommended.

\paragraph{Electrical networks.} Doyle--Snell,
\emph{Random Walks and Electric Networks} (1984, freely available
online) is a beautifully written elementary introduction. Lyons--Peres,
\emph{Probability on Trees and Networks} (Cambridge, 2016), is the
modern graduate reference. Spielman--Teng,
``Spectral sparsification of graphs'' (\emph{SIAM J.\ Comput.}, 2011),
introduces the spectral-sparsifier framework that motivates
nearly-linear-time Laplacian solvers (Cohen et al., FOCS 2014;
Kyng--Sachdeva, FOCS 2016).

\paragraph{Crystal graphs and the plactic monoid.} Kashiwara's
original paper, ``Crystalizing the $q$-analogue of universal
enveloping algebras'' (\emph{Comm.\ Math.\ Phys.}, 1990), established
the theory. Bump--Schilling, \emph{Crystal Bases: Representations
and Combinatorics} (World Sci., 2017), is the canonical book.
Hong--Kang, \emph{Introduction to Quantum Groups and Crystal Bases}
(AMS, 2002) covers the representation-theoretic background. For the
plactic monoid: Lascoux--Sch\"utzenberger, ``Le mono\"\i de plaxique''
(\emph{Quad.\ Ricerca Sci.}, 1981); Fulton, \emph{Young Tableaux}
(LMS Student Texts, 1997).

\paragraph{Graph isomorphism and complexity (broader).}
K\"obler--Sch\"oning--Tor\'an, \emph{The Graph Isomorphism Problem:
Its Structural Complexity} (Birkh\"auser, 1993), is a comprehensive
treatment. Babai's ``Graph isomorphism: the emerging paradigm'' (ICM,
2018) summarises the post-2015 algorithmic landscape.

\paragraph{General references.} Stanley's
\emph{Enumerative Combinatorics} (Vols.\ I and II, 2nd eds.) covers
much of the underlying combinatorial machinery. Lov\'asz's
\emph{Combinatorial Problems and Exercises} (2nd ed., 1993) is a
classical problem book. The arXiv \texttt{math.CO} archive contains
recent research; for graduate-level surveys see the
\emph{S\'eminaire Lotharingien de Combinatoire} archive.
